Zasp52’s differentially expressed intrinsically disordered region confers thin filament stability at the Z-disc

Version 1

Abstract

The Drosophila scaffolding protein Zasp52 is required to maintain structure at the muscle Z-disc which experiences strong forces during contraction. It is alternatively spliced into many isoforms, some of which contain a long intrinsically disordered region (IDR). We show that this region is primarily expressed in the indirect flight muscle (IFM) and is required for maintaining integrity of the Z-disc. Deleting the IDR-encoding exon15e results in flightlessness and structural IFM defects, including sarcomere bending at the Z-disc and an inability to de-contract. These defects are indicative of a lack of proper thin filament anchoring to the Z-disc. This is further supported by a genetic interaction between exon15e and actin. Fluorescence recovery after photobleaching of an isoform lacking exon15e shows that the IDR is required for maintaining Zasp52 at the Z-disc and thereby stabilizing Z-discs. Lastly, we can rescue these phenotypes by restricting IFM use. Together, these results suggest that Zasp52’s IDR confers thin filament stability at the Z-disc of IFM.

Background

The fruit fly Drosophila melanogaster possesses 152 muscle groups in the imago, as well as 54 in the larva (Bothe & Baylies, 2016). Each muscle is adapted to perform a specialized function, such as the extremely ordered indirect flight muscle (IFM) which is designed so as to enable the rapid oscillatory contractions needed to sustain flight. The marked phenotypic differences among these muscle types begs a molecular explanation. Alternative splicing is known to contribute to muscle type development and function with about 80% of sarcomere proteins being differentially expressed or spliced in Drosophila muscle (Nikonova et al., 2020). One such protein is Zasp52, a scaffolder, which contains numerous alternatively spliced exons generating a variety of isoforms.

All metazoan muscles share considerable structural homology (Steinmetz et al., 2012). They come in two types: striated, which appear striped given the presence of sarcomeres, and smooth, which do not possess sarcomeres. The sarcomere, a cylindrical structure, is the fundamental contractile unit of striated muscle and is bordered on either side by a dense circular structure called the Z-disc. Sarcomeres are arranged end-to-end in long arrays called myofibrils. Thin filaments composed of actin are anchored at and extend perpendicularly from the Z-disc into neighbouring sarcomeres, terminating at the H-zone (the region halfway between two Z-discs with no thin filaments). Thick filaments composed of muscle myosin are centred at the M-line (found at the centre of the H-zone) and interdigitate with thin filaments in a hexagonal lattice. The mechanism of sarcomere contraction is explained by the sliding filament theory in which thick filaments pull upon thin filaments in an ATP-dependent manner, bringing the Z-discs closer together. The simultaneous contraction of tandem sarcomeres comprising a myofibril is what leads to large-scale muscular contraction (Krans, 2010). The Z-disc is crucial for maintaining mechanical stability of the muscle and contains numerous proteins including Zasp52 which interacts with the thin filament cross-linker alpha-actinin (Liao et al., 2016) and actin (Liao et al., 2020) in Drosophila.

Zasp52, part of the three-membered Drosophila Alp/Enigma protein family along with Zasp66 and Zasp67, is crucial for maintaining structure at the Z-disc. It localizes to the Z-disc in all muscle types of Drosophila melanogaster (Katzemich et al., 2011) and is required for the assembly and maintenance of IFM sarcomeres (Katzemich et al., 2013; Liao et al., 2016). Developmentally, Zasp52 appears in the embryo first at myotendinous junctions and also at Z-bodies - clusters which later coalesce to form Z-discs (Katzemich et al., 2013). Zasp52 binds actin via its N-terminal PDZ domain plus a long C-terminal extension (Liao et al., 2020), and α-actinin via a region encompassing the PDZ domain plus a short C-terminal extension (Liao et al., 2016). In addition to the PDZ and Zasp motif (ZM) domain, Zasp52 contains four LIM domains which interact with the ZM domain resulting in auto-oligomerization; this mechanism is believed to drive radial growth of the Z-disc as well as pathological aggregate formation (González-Morales et al., 2019). Its primary human ortholog LDB3/ZASP is implicated in muscular dystrophies which are characterized by progressive adult-onset distal myofibrillar myopathy and sometimes cardiomyopathy (Griggs, et al., 2007).

Zasp52 is alternatively spliced into twenty-two different isoforms (according to FlyBase release FB2025_02, used throughout this paper), which leads to a variety of different domain and inter-domain combinations. The inter-domain linker regions connecting Zasp52’s structured domains (i.e., PDZ, ZM, and LIM domains) are less characterized but play an important role in protein function (Fisher & Schöck, 2022). Isoforms containing the first and second LIM domains contain a linker region between the two which can vary greatly in size. Encoded by several exons, this linker region may contain exon 15 which undergoes alternative 5’ splice site selection generating five exonic variants (15a-e) ranging from 130 nucleotides (15a) to 4429 nucleotides (15e) in length (Fig. 1A). The last 29 nucleotides of exon 15 contribute to the second LIM domain, but the remainder encodes the linker. When exon 15e is included, it constitutes the majority of the entire protein, encoding 1475 amino acids which total 160 kDa. In the canonical and longest isoform Zasp52-PF, the entire protein is 2194 amino acids long and the linker region between LIM1 and LIM2 is 1686 amino acids long. Exon 15e is referred to variously by other sources: as exon 9 in Flybase, exon 16c in Katzemich et al., 2011, and exon 18/19 in Benna et al., 2009; protein isoforms containing exon 15e were previously also referred to as Z(210) (Chechenova et al., 2013).

exon 15e-containing isoforms are expressed in a specific spatiotemporal manner

(A) A schematic of the Zasp52 locus. Exons are drawn to scale, introns are not. Coding exons are orange and untranslated regions in grey. Alternative start sites are also indicated. Below in green are mappings of structured domains according to SMART (Letunic & Bork, 2018); and also Katzemich et al, 2011 which defined LIM1 and LIM2 variants. The deletion in ex15e mutants is indicated in red. (B) Western blot using five different tissue extracts from control flies probed with a Zasp52 full-length antibody. Relative amounts of Zasp52 content as measured in each lane, over the total protein content as measured using the TGX Stain-free gel method, are quantified below for each tissue. Exon 15e content as measured using the protein content above the dotted blue line for each lane, over Zasp52 content is quantified below that. Molecular weights appear higher, perhaps due to retardation caused by non-specific interactions with the gel matrix, as also seen in Watts et al., 2017, Fig. S1. (C) RNA-seq data from Spletter et al., 2018 of IFM tissue across various timepoints after puparium formation; the last timepoint is of adults one day after eclosion. Gene expression levels in transcripts per million (TPM) are shown for Zasp52 isoforms classified according to their splicing of exon 15.

Though not much is known about this linker region between LIM1 and LIM2, its predominant expression in IFM and unusually large length prompted us to investigate it. We were curious to see how deleting only exon 15e while retaining all conserved domains would affect muscle structure and function. This linker was found to be highly disordered; intrinsically disordered regions (IDRs) have mostly been analyzed biochemically, with only limited data on their in vivo function in model organisms (Jensen et al., 2025). Flight ability was impaired and myofibrils displayed a variety of phenotypes which were all indicative of a loss of thin filament anchoring at the Z-disc. Most phenotypes exhibited an age-related increase in severity. Put together, these data demonstrate a crucial role of this large IDR in maintaining sarcomere integrity by stabilizing thin filaments at the Z-disc in IFM.

Results

exon 15e-containing isoforms are expressed in a specific spatiotemporal manner

Western blotting using a polyclonal antibody capable of detecting all Zasp52 splice variants (hereinafter referred to as the Zasp52 full-length antibody, Jani & Schöck, 2007) was used to analyze the presence of Zasp52 in IFM, ovaries, whole larvae, heads, and whole male abdomens (Fig. 1B). IFM contains the second most Zasp52 protein of any isoform relative to total protein content compared to the other tissues analyzed and is the only tissue that significantly contains exon 15e-containing isoforms relative to total Zasp52 content. This agrees with previous studies (Chechenova et al., 2013, Benna et al., 2009) which found no higher weight isoforms in larvae, and with transcriptomics data. Whole organism RNA-seq data from the ENCODE project (Duff et al., 2015) reveals very little expression of exon 15e in all developmental stages up until four days past the white prepupal stage (Fig. S1). To verify that this is indeed a phenomenon of exon 15e inclusion and not merely Zasp52 expression, one can look at the surrounding exons which do not follow the same pattern as exon 15e. To gain a more precise perspective on IFM expression of exon 15e, we turned to an RNA-seq dataset from Spletter et al., 2018 which looked at total RNA content of IFM at various developmental time points (Fig. 1C). Expression of all Zasp52 transcripts in IFM surges from 48 to 72 h after puparium formation (APF), with most of these transcripts containing exon 15e, as opposed to including exon15 a-d only or no exon 15 at all.

We used various software to predict the level of disorder across Zasp52-PF (Fig. S2A). All identified the linker region between LIM1 and LIM2 as an intrinsically disordered region (IDR) to various degrees, although some regions within it appear more ordered. The actin-binding motif identified by Ashour et al., 2023 within exon 15e is predicted to be more structured than the surrounding regions. No significant sequence similarity was found when aligning Zasp52’s linker region to human or other Drosophila melanogaster ALP/Enigma protein family members. We then compared the lengths of the linker regions between LIM1 and LIM2 in all insect species possessing a Zasp52 ortholog as defined by NCBI’s Eukaryotic Genome Annotation Pipeline (Fig. S2B). Drosophilids generally have a longer linker than their taxonomical neighbours. Mayflies, dragonflies, and damselflies, which use direct flight muscle, have notably short linkers. Since linear sequence often does not fully describe an IDR’s function, we analyzed linear clustering or mixing of different residue types using NARDINI (Cohan et al., 2022). The z-score matrix of the linker region between LIM1 and LIM2 in Zasp52-PF showed a strong clustering/segregation of polar residues (μ), as well as alanine (A) and proline (P), and hydrophobic residues (h) (Fig. S2C). The z-score matrix of exon 15e alone displayed much of the same clustering, except with additional strong clustering of acidic residues (-) with respect to other residue classes (Fig. S2C’). This implies that the additional exons in the linker region dilute or mask the strong clustering of acidic residues in exon 15e alone. We attempted to compare z-score matrices across various arthropod species but were unable to detect any meaningful correlation indicating that the binary patterns uncovered by NARDINI do not reveal taxonomical differences in exon 15e composition.

CRISPR deletion of exon 15e displays flight defects

To further investigate the function of exon 15e, we generated a CRISPR deletion spanning five intronic nucleotides five prime to exon 15e up to 149 nucleotides before the 3’ end of exon 15e (Fig. 1A). In total, 4285 nucleotides totaling 156 kDa were deleted. Thus, our mutants would affect all isoforms containing exons 15b-e. The deletion endpoints were sequenced to confirm effective CRISPR knockout. To confirm that our deletion worked and spliced as intended, we performed Western blots using the Zasp52 full-length antibody to compare our deletion with a control and the MiMIC line MI00979 (Venken et al., 2011) which truncates everything C-terminal to and including exon 15e (Fig. 2A). The largest protein that could possibly be present in our mutant would be 136 kDa (Zasp52-PI). Our deletion was devoid of all higher molecular weight isoforms containing exon 15e which were present in the control as expected. Our deletion also contained several isoforms which were heavier than the highest molecular weight isoforms of MI00979 which were also not present in control, suggesting that these are novel products containing the C-terminal LIM domains. To confirm that this was indeed the case, i.e., that the three C-terminal LIM domains were retained, we raised an antibody against them and performed a Western blot (Fig. 2A’). Our CRISPR deletion retained these LIM domains as expected. We refer to this CRISPR mutant as ex15e.

A CRISPR deletion of exon 15e displays flight defects

Western blots using either the Zasp52 full-length antibody (A) or antibody against LIM2-LIM4 (A’) of either w1118 (ctrl hereinafter), ex15e, or the MiMIC line MI00979 whole flies. Relative amounts of Zasp52 content as measured in each lane, over the total protein content, are quantified below for each genotype. (B) Larval crawling assay of either ctrl (n=21) or ex15e (n=21) third instar larvae showing no significant difference between the two (Unpaired t-test with Welch’s correction; p=0.6006). (C) Flight assays of ctrl vs ex15e flies of three different age categories: one-day-old (ctrl n=138 flies, ex15e n=123), five-days-old (ctrl n=125, ex15e n=119), and three-weeks-old (ctrl n=86, ex15e n=81). The y-axis indicates flight strength: flies were released into a tube and those that landed in the top segment (y=1) had the strongest flight strength while those that landed in the bottom dish (y=8) had the weakest. The x-axis indicates the proportion of flies that landed in that segment. Red arrows indicate the average position landed in control flies, yellow arrows are for ex15e flies. Difference in flight ability was statistically significant for all age categories (Fisher’s exact test; p<0.0001).

We wanted to see how ex15e affected the locomotory abilities of the fly. Since larvae do not express exon 15e, we do not expect them to be affected by our deletion. Therefore, we performed a larval crawling assay as a negative control to confirm that ex15e larval muscles were unaffected (Fig. 2B). As predicted, control and ex15e larvae demonstrated no differences in crawling ability. Next, we wanted to see how flight would be affected, given that exon 15e is present primarily in the adult IFM. ex15e flies of all age categories had impaired flight ability compared to control with a severe deterioration at three weeks of age (Fig. 2C). To ensure that no secondary-site mutations were affecting our mutant phenotypes, we performed a simple binary flight assay using the heterozygous ex15e deletion over a Zasp52 deficiency (Fig. S3A).

This impairment in flight could stem from a variety of causes, given the multimodal functionality of Zasp52. We decided to first look at gross muscle morphology before zooming in on our investigation. Thorax bisections revealed no noticeable difference in the IFMs of ex15e and control flies (Fig. S3B, C). Since Zasp52 was known to be involved in integrin interactions (Jani & Schöck, 2007) and junctional components (Ashour et al., 2023), we inspected the myotendinous junctions of IFM, which are called modified terminal Z-discs (MTZ), in three-week-old flies (Fig. S3D-G’’). βPS integrin localization appeared unaffected in ex15e; this agrees with integrins being upstream of Zasp52 (Jani & Schöck, 2007). However, the MTZ appeared disrupted and malformed, lacking the organized structure seen in controls. Furthermore, we also noticed myofibril defects which prompted us to further investigate IFM at the myofibrillar level.

ex15e mutant IFM phenotypes: bending at the Z-disc and thin filament intrusion into the H-zone

IFM tissue was analyzed by confocal microscopy with several phenotypes being observed. First, myofibrils in ex15e mutants were often found to be bent at the Z-disc (Fig. 3A-C). In control flies, sarcomeres across different age groups typically displayed the same proportion of degree of bentness. In ex15e, the proportion of bent sarcomeres increased with age, as well as the degree of bending, compared to control flies. Force-induced bending results in a sarcomere whose inner curve is bowed, much like a hose with a kink. The bending in ex15e appears to be a result of the improper anchoring of thin filaments which leads to a shear type angle shift in the sarcomere at the Z-disc. Bending was found exclusively at the Z-disc and not the M-line in all muscle analyzed.

ex15e mutant IFM phenotypes: bending at the Z-disc and thin filament intrusion into the H-zone

(A-C) Analysis of bent sarcomere phenotype. Actin is visualized by phalloidin in green. Control myofibrils (A) appear straight while ex15e (B) myofibrils contain a range of bending (representative three-week-old myofibrils shown). Red arrow indicates a representative sarcomere with a bend of <145°, yellow indicates 145-165°, and white >165°. Bend angles were statistically equivalent (≡) between ctrl and ex15e in one-day-olds and five-day-olds, but not equivalent (ne) in three-week-olds, where ex15e displayed greater bentness (C). Sample sizes: one-day-old (ctrl n=1242 sarcomeres, ex15e n=1851), five-days-old (ctrl n=1192, ex15e n=946), and three-weeks-old (ctrl n=1768, ex15e n=1125) (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; 1d.o. tL=12.95, tU=-12.07; 5d.o. tL=6.03, tU=-8.01; 3w.o. tL=-6.34, tU=-20.63). (D-F) Sarcomeres also exhibited actin in the H-zone phenotype (representative three-week-old myofibrils shown). Control (D) myofibrils display normal H-zones without actin with M-lines/H-zones indicated by blue arrows. ex15e (E) myofibrils display actin accumulation at the H-zone; arrows indicate same as above. Actin is visualized by phalloidin in green and Z-discs are identified by the full-length Zasp52 antibody in red. These sarcomeres with H-zone actin, which appear hypercontracted, are significantly overrepresented in ex15e compared to control across all age categories (F). Sample size were as follows: one-day-old (ctrl n=3255 sarcomeres, ex15e n=4056), five-days-old (ctrl n=3221, ex15e n=2677), and three-weeks-old (ctrl n=5394, ex15e n=4720 (Fisher’s exact test; p<0.0001). To confirm that H-zone actin sarcomeres were indeed contracted, the Z-to Z-disc length was measured in ex15e three-week-old flies of both sarcomeres with a bare H-zone (H-zone normal, n=89 sarcomeres) and those with actin in the H-zone (H-zone actin, n=117) (G). H-zone actin sarcomeres were significantly smaller (Unpaired t-test with Welch’s correction; p<0.0001).

Another phenotype observed in ex15e was the presence of a dense band of actin staining at the H-zone, where normally there would be no actin staining (Fig. 3D-G). Control flies have very few sarcomeres with this phenotype at all ages; ex15e are similar to control at one day of age but begin to accumulate a large number of these abnormal sarcomeres with time. At three weeks of age, a large proportion of sarcomeres have actin at the H-zone. Usually, groups of myofibrils in the same area would exhibit this phenotype rather than these abnormal myofibrils being randomly dispersed among the normal ones. This dense band of actin suggested that thin filaments were overlapping at the M-line. This could be a result of hypercontraction, where thin filaments are brought so close together that they overlap. However, the protocol used for these preparations includes a relaxing solution containing ATP, which induces cross-bridge detachment resulting in the adoption of a relaxed conformation. To see if these abnormal sarcomeres with actin in the H-zone were indeed contracted, we measured their length and compared them to normal sarcomeres from the same sample in three-week-old flies (Fig. 3G). Sarcomeres with actin in the H-zone had a smaller length than those without indicating that they were indeed contracted. Muscle decontraction relies on the elastic nature of the sarcomere to restore it to a relaxed conformation once myosin heads detach from the thin filaments. This requires stable Z-discs and myofilament attachments, the lack of which may be the cause of a permanent hypercontracted state in ex15e.

To gain a better understanding of this phenomenon, we performed transmission electron microscopy (TEM) of myofibrils and noticed phenotypes that concurred with our immunofluorescence confocal images (Fig. 4). Z-discs were highly disrupted and contained many holes and broken regions. M-lines were also affected in a similar manner. Gaps between filaments were abundant and in general appeared more disorganized than control. Filaments were often frayed at the edges of the sarcomeres.

TEM reveals defects in mutant sarcomere ultrastructure

TEM images reveal similar phenotypes as immunofluorescent ones (three-week-old shown). Control (A) sarcomeres are very structured and even. Longitudinal sections of ex15e (B,C) display broken Z-discs (red arrow), broken M-lines (yellow), bending at the Z-discs (green), splits in filaments (purple), and fraying peripheral filaments (blue). Transverse sections of ex15e (D) also display some of these abnormalities.

Most ex15e defects are restored by rescue constructs

To determine if the phenotypes are truly due to the IDR or are caused by an altered isoform composition, we wanted to see if the phenotypes of ex15e could be rescued. First, we expressed either UAS-Zasp52-PF or UAS-Zasp52-PR using a UH3-GAL4 driver in a homozygous ex15e background and examined three-week-old flies (Fig. 5B-H). Zasp52-PF is the canonical full-length Zasp52 isoform which contains exon 15e, while Zasp52-PR contains all structured domains but not exon 15e (Fig. 5A). The degree of bending in Zasp52-PR rescue sarcomeres was statistically equivalent to that of ex15e mutants, while that of Zasp52-PF rescues was equivalent to control flies, indicating that Zasp52-PF but not Zasp52-PR can rescue this phenotype (Fig. 5F). The proportion of H-zone actin (i.e., hypercontracted) sarcomeres was not significantly different between ex15e and Zasp52-PR rescues as expected. Interestingly, no hypercontracted sarcomeres were observed upon re-expression of Zasp52-PF again indicating a successful rescue (Fig. 5G). Finally, we tested flight ability and observed a significant increase in flight strength in Zasp52-PF rescues compared to Zasp52-PR rescues, with the latter being not significantly different from ex15e (Fig. 5H). Zasp52-PF rescues were still weaker fliers than control flies suggesting a partial rescue of flight ability.

Certain ex15e defects are restored by rescue constructs

Zasp52-PF or Zasp52-PR was heterozygously overexpressed with a UH3-GAL4 driver over a homozygous ex15e background in a rescue attempt. In this figure, all myofibrils had actin visualized with phalloidin in green, all experiments were done at three weeks of age, and ctrl and ex15e data are same as previous. (A) Schematic showing Zasp52-PF which is the canonical full-length isoform, and Zasp52-PR which contains all structured domains but lacks exon 15e. PDZ domain (blue), ZM (red), LIM domains (green). (B-E) Myofibrils display bending, Z-disc disruption, and overall myofibrillar disorganization in Zasp52-PR overexpressions (D), while Zasp52-PF overexpression (E) appears similar to ctrl (B). Bend angles were equivalent (≡) between ex15e and ex15e; UH3>Zasp52-PR, and between ctrl and ex15e; UH3>Zasp52-PF (F). Sample sizes: ex15e; UH3>Zasp52-PR n=2049 sarcomeres, ex15e; UH3>Zasp52-PF n=2221. (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; ex15e-ex15e; UH3>Zasp52-PR tL=1.92, tU=-10.16; ctrl-ex15e; UH3>Zasp52-PF tL=17.65, tU=-8.51). The proportion of sarcomeres that were contracted (i.e., actin in H-zone) was not significantly different between ex15e and ex15e-ex15e; UH3>Zasp52-PR, whereas no such sarcomeres were found in ex15e; UH3>Zasp52-PF (G). Sample sizes: ex15e; UH3>Zasp52-PR n=2843 sarcomeres, ex15e; UH3>Zasp52-PF n=2480. (Fisher’s exact test p=0.0521). (H) Flight ability between ex15e; UH3>Zasp52-PR (n=66) and ex15e; UH3>Zasp52-PF (n=85) was significantly improved, with ex15e; UH3>Zasp52-PR showing no significant difference from ex15e but ex15e; UH3>Zasp52-PF showing significant difference from ctrl. The y-axis indicates flight strength: flies were released into a tube and those that landed in the top segment (y=1) had the strongest flight strength while those that landed in the bottom dish (y=8) had the weakest. The x-axis indicates the proportion of flies that landed in that segment. Red arrows indicate the average position landed in control flies, yellow are ex15e, green are ex15e; UH3>Zasp52-PR, and blue are ex15e; UH3>Zasp52-PF. (Fisher’s exact test; ex15e; UH3>Zasp52-PR-ex15e; UH3>Zasp52-PF p<0.0001; ex15e-ex15e; UH3>Zasp52-PR p=0.2668; ctrl-ex15e; UH3>Zasp52-PF p=0.0129).

ex15e and actin genetically interact

We next wanted to test for a genetic interaction between exon 15e and Act88F (the actin isoform present in IFM) by analyzing heterozygous ex15e, heterozygous Act88FKM88 (an amorphic mutant of Act88F), and a di-heterozygous combination of both (ex15e/+; Act88FKM88/+) (Fig. 6). ex15e/+ resembled control myofibrils and Act88FKM88/+ has mild myofibrillar defects. ex15e/+; Act88FKM88/+ exhibited supra-additive effects where we often noticed a novel phenotype of very large H-zones. A ratio of twice the thin filament length to the total sarcomere length was computed for all genotypes. Lower values represented sarcomeres whose thin filaments occupied a smaller proportion of the sarcomere and consequently had larger H-zones, whereas a value of one represented the absence of an H-zone. This ratio was not significantly different between control, ex15e/+, and Act88FKM88/+, but was considerably smaller in di-heterozygotes (Fig. 6E), indicating a strong genetic interaction between Act88F and exon 15e.

Large H-zones demonstrate a genetic interaction between exon 15e and Act88F

(A-D) Myofibrils of one-day-old flies of various genotypes. Actin was visualized with phalloidin in green. Control and heterozygous ex15e myofibrils appear indifferentiable; heterozygous Act88FKM88 are slightly disrupted with fraying and slight narrowing. Representative ex15e/+; Act88FKM88/+ myofibrils with large H-zones are shown. The ratio of twice the thin filament length denoted distance “A” to the corresponding sarcomere’s length denoted as “B” for all four genotypes is quantified in (E). A schematic illustrates the measurement of these distances. Sample sizes were as follows: ctrl n=119 sarcomeres, ex15e/+ n=122, Act88FKM88/+ n=119, ex15e/+; Act88FKM88/+ n=119. The ratio A/B represents the proportion of the sarcomere length occupied by thin filaments and is inversely correlated with H-zone length; ex15e/+; Act88FKM88/+ sarcomeres have significantly larger H-zones than all others. All sarcomeres with an Act88FKM88 allele were radially narrower than control, but a synthetic enhancement could not be concluded from this observation. (Unpaired t-test with Welch’s correction; ctrl-ex15e/+ p=0.0502; ctrl-Act88FKM88/+ p=0.2770; ctrl-ex15e/+; Act88FKM88/+ p<0.0001; ex15e/+-Act88FKM88 p=0.4938; ex15e/+-ex15e/+; Act88FKM88 p<0.0001; Act88FKM88/+-ex15e/+; Act88FKM88 p<0.0001).

FRAP reveals defects in ex15e protein dynamics

Given the differences we have thus observed in our mutants, as well as the great length of the IDR, we decided to investigate Zasp52 kinetics in ex15e sarcomeres by performing Fluorescence Recovery After Photobleaching (FRAP) (Fig. 7A, C). We expressed GFP-tagged Zasp52-PR (see Fig. 5A) which lacks exon 15e in either a wild-type background as our control (UH3>GFPZasp52-PR), or in an ex15e background (ex15e; UH3>GFPZasp52-PR). We demonstrate in Fig. 5 that Zasp52-PR cannot rescue ex15e. One-day-old IFM Z-discs were photobleached and recovery of fluorescence was monitored. Fluorescence recovery in the ex15e background was to a significantly higher level indicating a larger mobile fraction (Fig. 7B). A greater mobile fraction in our mutant background signifies that exon 15e is required to hold Zasp52 in place at the Z-disc, either by fastening itself, or keeping other molecules in place thereby contributing to overall Z-disc stability.

FRAP reveals defects in ex15e protein dynamics

FRAP experiments visualizing recovery of GFPZasp52-PR were performed on one-day-old female IFM Z-discs either in our mutant background (ex15e;UH3>GFPZasp52-PR, n=9 flies) or wild-type background which served as a control (UH3>GFPZasp52-PR, n=12) (A). Control UH3>GFPZasp52-PR Z-discs fixed with 4% paraformaldehyde for 10 minutes (labelled “Fixed”, n=7) displayed no recovery as expected. Plateaus of fitted curves were significantly different, with a greater mobile fraction in the mutant background (Unpaired t-test with Welch’s correction; p<0.0001) (B). Representative snapshots of bleached Z-disc recovery illustrate this difference between genotypes (C).

Most ex15e defects are rescued by immobilization

On a different note, we noticed that virtually all the ex15e phenotypes worsened with age, which suggests that use might contribute to deterioration. We therefore decided to prevent flies from using their IFMs by immobilizing them (Fig. 8). Flies were reared until three weeks of age in between glass plates separated by 1.5 mm spacers which prevented them from flying or even attempting to fly. The degree of bending in immobilized flies was equivalent to that of control flies (Fig. 8D). The proportion of hypercontracted sarcomeres was also not significantly different between immobilized and control flies (Fig. 8E). Intriguingly, even flight ability was restored in immobilized flies compared to age-matched controls with no significant difference between the two (Fig. 8F). These results suggest successful rescue by preventing IFM use via immobilization.

Certain ex15e defects are rescued by immobilization

Rescues were attempted by immobilizing flies thereby preventing IFM use. ex15e myofibrils in flies that were immobilized (C) resembled those of ctrl (A). Bend angles were equivalent between ctrl and ex15e immobilized flies (n=2119) (D). (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; tL=15.36, tU=-12.00). The proportion of sarcomeres that were contracted was not significantly different between ctrl and immobilized flies (n=2246) (E). (Fisher’s exact test p=0.2026). Flight ability was rescued with immobilized flies (n=67 flies) having no significant flight defect compared to control (n=86) but being significantly improved from their non-immobilized counterparts (n=81) (F). (Fisher’s exact test; ctrl-immobilized p=0.0640; ctrl-ex15e and ex15e-immobilized p<0.0001).

Discussion

Traditionally, IDRs were associated with the concept of amorphousness, finding themselves in transient structures such as condensates and membraneless organelles. As our understanding of disordered proteins has changed, their role in more disparate contexts has become increasingly appreciated, revealing them as modular hubs, flexible scaffolds, and more (Cortese et al., 2008). Furthermore, their role in disease has spurned great interest in their myriad functions. In this study, we demonstrate the role of an exceptionally large disordered region, Zasp52’s exon 15e, in maintaining stability and integrity in a highly rigid and force-bearing structure – the Z-disc.

First, we showed that the expression of exon 15e is spatiotemporally regulated. Being confined to the adult IFM suggests that exon 15e’s function is uniquely required for the rapid oscillatory contractions in indirect flight. Supporting this is the fact that the linker region containing exon 15e is generally much longer in Drosophilids, and rather short in insects using direct flight. Although this linker region is predicted by multiple software programs to be disordered, within it are regions of greater order, one of which corresponds to a putative actin-binding motif (Ashour et al., 2023). This arrangement exists in other proteins such as Cno in Drosophila, where higher-ordered actin-binding regions are embedded in a large IDR (Jensen et al., 2025). Furthermore, a synthetic enhancement between ex15e and Act88FKM88 heterozygotes indicates that exon 15e and actin genetically interact. There may be other important short linear motifs (SLiMs) contained within the linker region which serve as interacting sites with other sarcomere proteins. Indeed, multiple sequence alignment of the linker regions of various insects reveals several conserved motifs of unknown function. The presence of an actin-binding motif that may strengthen thin filament anchoring to the Z-disc would be in line with the phenotypes we observed in ex15e. Bending at the Z-disc is indicative of a loss of thin filament anchoring and stability at the Z-disc, allowing them to shift out of alignment. Normal sarcomeres bent due to force would exhibit a kinked-hose appearance, where the inner edge is sharply bent and the outer one bowed. This was occasionally observed in controls, whereas ex15e bends were always shifted uniformly. Actin in the H-zone is also indicative of loss of stable thin-filament anchoring. During preparation, IFMs are incubated overnight in skinning solution during which they enter a state of rigor. After washing them with an ATP-containing relaxing solution, the myosin heads detach from the thin filaments and the supple nature of the sarcomere which provides it with elastic recoil, causes it to resume a relaxed conformation. However, when thin filaments are weakly anchored at the Z-discs, the natural tension contained within the sarcomere is reduced and thus it will not “bounce back” into a relaxed conformation, resulting in a hypercontracted state which manifests as actin continuing to reside in the H-zone. Cheerio, a Drosophila filamin, was shown to stably attach actin thin filaments at the Z-disc and induces a similar actin incorporation into the H-zone when mutated (González-Morales et al., 2017). Thus, we also demonstrate the importance of Z-disc integrity in allowing sarcomeres to return to their relaxed state in rapidly contracting asynchronous IFM. Accordingly, synchronous muscles do not have such regularly structured Z-discs, and do not require Zasp52’s exon 15e. Indeed, exon 15e is absent in larval and other muscle types. We therefore propose that exon 15e is uniquely required to stabilize rapidly contracting asynchronous IFM.

The orthologous human ALP/Enigma family consists of seven members, with LDB3 (ZASP) considered the canonical Zasp52 ortholog. Zasp52 is more precisely a combined ALP/Enigma member; its first LIM domain is sequentially homologous to the single LIM domain of human ALP family members, and its last three LIM domains are homologous to the three Enigma family LIM domains. Though these human proteins do contain disordered linkers connecting their structured domains, they are much shorter than in Drosophila and are not sequentially homologous. The length of Drosophila’s linker was likely evolutionarily driven to be large, as we see insects using direct flight to have much shorter linkers. Several pathological mutations have been identified in the linker region between LDB3’s PDZ and LIM domains (Vatta et al., 2003), some of which are now known to lie in the actin-binding Zasp motif (ZM) (Lin et al., 2014). Interestingly, LDB3 has three skeletal muscle-specific splice isoforms which are differentially expressed and variably retain/omit exons 9 through 11 of a linker sequence (Lin et al., 2014). The prenatal long (L) isoform contains exons 10 and 11, post-natal long (LΔex10) has only exon 11, and short (S) has only 9. Post-natal LDB3-LΔex10 contains an actin-binding region in its linker region whose function is abolished by the inclusion of exon 10 in LDB3-L. Thus, human Zasp protein linkers also function in actin binding, regulated by developmentally differential expression. Linker regions in several human ALP/Enigma proteins were found to interact with both α-actinin and actin using AlphaFold (Healy & Collins, 2023). The mouse ortholog Cypher also contains tissue-specific splice variants which variably include or omit linker sequences (Huang et al., 2003). Like Zasp52’s exon 15e, Cypher is not required for sarcomerogenesis or Z-disc assembly but only maintenance (Zhou et al., 2001).

Our genetic rescues support our hypothesis of exon 15e involvement in stabilizing thin filaments and Z-discs. The rescues also demonstrate that the conserved domains (PDZ, ZM, LIM1-4) are not sufficient for proper Z-disc and thin filament stability in IFM, because Zasp52-PR could not rescue any of the defects. Interestingly, these phenotypes, as well as the flight defects we observed, worsened with age, with defects in one-day-old flies being very minor. This implies that exon 15e is not required developmentally, but rather for the maintenance of sarcomeres which naturally deteriorate with use. We originally hypothesized that since IDRs often lead to protein aggregation and phase separation, exon 15e might drive the formation of Z-bodies, primitive Z-discs which form through the coalescence of diffuse Z-disc proteins including Zasp52. However, given the lack of defects in early Z-discs and the fact that exon 15e expression does not temporally coincide with Z-body formation, we reject this hypothesis. The role of exon 15e in maintenance is supported by immobilization rescues, in which flies were prevented from flying. We opted for a low-ceiling apparatus which we believe would not only prevent flies from flying but discourage them from futilely attempting to utilize their IFM at all. Surprisingly, immobilization restored flight ability to wild type levels. Interestingly, human zaspopathies often manifest late in life, with one study citing onset at 44 to 73 years of age (Selcen & Engel, 2005), much like what we observed in flies. Our data suggest that avoiding strenuous exercise might delay the onset of zaspopathies. Given the maintenance-specific role of exon 15e, but also that as a whole, Zasp52 is indispensable for pupal myofibril development and not just maintenance (Katzemich et al., 2013), Zasp52 appears to have a multifaceted role more complex than previously considered.

Although one-day-old ex15e myofibrils are morphologically indistinguishable from controls, we were already able to detect abnormalities in Zasp52 dynamics using FRAP. The presence of recovery suggests that Zasp52 is not immobile as might be assumed for an anchoring protein and does indeed possess kinetics on a sub-minute timeframe. This is supported by significant transcription of the gene, including exon 15e-containing isoforms, into adulthood (Fig. S1). The mobile fraction of Zasp52-PR was significantly increased in a system entirely lacking exon 15e (the PR isoform also lacks exon 15e) indicating a role of exon 15e in retaining Zasp52 at the Z-disc, whether by direct interaction or not. The FRAP data suggest that Zasp52’s IDR stabilizes the Z-disc, in line with the electron microscopy data. It suggests that Z-discs can be described as highly solid aggregates in part stabilized by IDRs, on the other end of the spectrum of the transient condensates initially described, thereby further expanding the importance of IDRs for establishing subcellular structures.

Overall, Zasp52 and its isoforms represent a complex toolbox of multifunctional proteins. There has been evidence suggesting that Zasp52 is not required for initial sarcomere assembly in embryos (Rui et al., 2010), important for Z-disc maintenance overall (Rui et al., 2010, Jani & Schöck, 2007, Katzemich et al., 2013), and required for pupal myofibril assembly (Katzemich et al., 2013). The complete picture of Zasp52’s varied function is still unclear; however, we have shown that a specific disordered region is required for maintaining IFM and stabilizing the rigidity of Z-discs. The long linker might not only contain protein-interacting regions but also provide the spacing needed for the neighboring structured domains to properly interact with their partners and perform their roles. We have seen in our genetic interaction experiment an inability of thin filaments to properly elongate when Act88F is partially depleted, which further implicates exon 15e not only in Act88F structure but dynamics. Most encouraging is the complete rescue of flight ability through immobilization, which may provide avenues for delaying the onset of human zaspopathies through reducing strenuous exercise.

Materials and Methods

Fly strains and husbandry

Flies were housed at room temperature (22°C) in standard glass vials and food unless indicated otherwise. Immobilized flies were reared in between 3.25×3.25 inch glass plates separated by 1.5 mm spacers; standard food was inserted into the gap. ex15e flies were created by Genetivision using CRISPR-mediated homology-directed repair. The gRNA sequences used were TTTGAACTTGAACTCTAATGTGG for the downstream target and CAGCCTCTCCACCTTTCCGATGG for upstream. The deletion was replaced with a cassette containing a 3xP3DsRed marker flanked by attP landing sites. It was validated by sequencing at the cut sites. The following genotypes were used: ex15e, w1118 as control in all experiments, Df(2R)BSC427, Act88FKM88, UAS-GFPZasp52-PR, UH3-Gal4, UAS-Zasp52-PF, UAS-Zasp52-PR, Zasp52MI00979. The UAS constructs consist of the respective Zasp52 isoform with an N-terminal ALFA-tag and C-terminal Spot tag. UAS constructs were made by Genscript and injected into w1118 by GenomeProlab.

Western blotting

Extracts were prepared by homogenizing ten flies/fly parts in 100 μL RIPA buffer with protease inhibitor plus an equal amount of 2× Laemmli buffer with β-mercaptoethanol. Samples were boiled at 95°C for five minutes and centrifuged at 12,000×rcf for 5 minutes. Supernatant was used for electrophoresis according to standard protocol using a 4-15% BioRad TGX Stain-free gel. Stain-free gels were imaged for total protein content on a BioRad ChemiDoc before protein was transferred onto a nitrocellulose membrane. Standard Western blotting procedures were followed. Primary antibodies used were rabbit anti-Zasp52 full-length (Jani & Schöck, 2007), and a rabbit antibody against the three C-terminal LIM domains of Zasp52 (FEKYLAPTCSKCAGKIKGDCLNAIGKHFHPECFTCGQCGKIFGNRPFFLEDGN AYCEADWNELFTTKCFACGFPVEAGDRWVEALNHNYHSQCFNCTFCKQNLEGQSFYNKGGRPF CKNHAR) made by Genscript; secondary antibody was goat anti-rabbit IgG (H+L) HRP (Invitrogen).

Locomotory assays

For flight assays, flies were reared in vials (unless immobilized) until the desired age, with vials flipped whenever pupae began to appear. Flight assays were conducted in a manner similar to that described in Babcock & Ganetzky, 2014. A tube 22 cm long and 3.2 cm in diameter was placed in a funnel; the bottom of the tube rested 4 cm above the bottom spout of the funnel. The funnel rested in a 60 cm long tube which was 13.5 cm in diameter and covered with TAD Shake & Spray (T-141) on the inside. The bottom of the funnel was 12 cm below the top of the larger tube. The larger tube was divided vertically into seven 8.5 cm segments, with segment 1 being at the top, and segment 7 being at the bottom. A dish containing water was placed below the tube and was considered segment 8. Vials containing flies were dropped down the narrow tube so that they would be ejected into the larger tube and land somewhere along its length. Flies naturally fly upwards and those with stronger flight would land higher up the tube. Binary flight assays were conducted by releasing flies one at a time and recording whether they could generate any upwards lift. They were raised in the same conditions as above.

For larval crawling assays, third instar larvae of the desired genotype were collected from bottles and assayed according to Post & Paululat, 2018. In brief, videos of larvae on a slightly moistened petri dish were taken. Time segments where larvae were crawling in a straight line continuously for at least ten seconds were analyzed – contractions per distance were measured and recorded.

Muscle preparation and immunofluorescence

Primary antibodies used were rabbit anti-Zasp52 full-length (Jani & Schöck, 2007), rabbit anti-LIM234 (this study), and mouse anti-βPS-integrin (CF.6G11; obtained from Developmental Studies Hybridoma Bank, Brower et al., 1984) at a concentration of 1:200 for all. Secondary antibodies used were Alexa Fluor 647 goat anti-rat (Life Technologies) and Alexa Fluor 488 goat anti-rabbit (Invitrogen) at a concentration of 1:400 for all.

For IFM tissue, a protocol similar to Xiao et al., 2017 was followed (refer to for recipes). Briefly, thoraces with bisected and incubated in Relaxing-Glycerol solution at -20°C overnight. IFM were isolated and fixed in 4% paraformaldehyde for 15 minutes. Fixed IFM were incubated in primary antibodies at 4°C overnight. IFM were incubated in Acti-stain 555 fluorescent phalloidin (Cytoskeleton Inc) and secondary antibodies, at room temperature for one hour. Samples were mounted in Mowiol mounting medium.

For thorax hemisections including MTZ, a scalpel was used to sagittally bisect thoraces. Samples were fixed in 4% paraformaldehyde for 15 minutes and incubated in primary antibodies for one hour at room temperature, and then Acti-stain 555 fluorescent phalloidin and secondary antibodies for one hour at room temperature. Samples were mounted in Mowiol mounting medium using three 0.12 mm spacers.

For TEM, bisected thoraces were treated with 5 mM MOPS, 150 mM KCl, 5 mM EGTA, 5 mM MgCl2, 5 mM ATP, 1% Triton X-100, and protease inhibitor for two hours. then the same solution with 50% glycerol instead of Triton X-100 overnight at 4°C. This was repeated at -20°C. Samples were washed in 5 mM MOPS, 40 mM KCl, 5 mM EGTA, 5 mM MgCl2, 5 mM NaN3, protease inhibitor, and then fixed in 3% glutaraldehyde and 0.2% tannic acid for two hours. Samples were washed in 20 mM MOPS, 5 mM EGTA, 5 mM MgCl2, 5 mM NaN3, and then in 100 mM sodium phosphate buffer, 5 mM MgCl2, 5 mM NaN3. Secondary fixation was performed in 1% osmium tetroxide in the previous sodium phosphate buffer for one hour. Samples were stained in 2% uranyl acetate, washed with water, and underwent a serial dehydration in ethanol. Samples were washed with propylene oxide and serially infiltrated with Mollenhauer Epon-Araldite. After curing in Beem capsules, 70 nm sections were cut and stained in 4% uranyl acetate followed by SATO lead stain.

Fluorescence Recovery After Photobleaching

One-day-old female flies were anaesthetized and their abdomens, wings, and legs removed. Carcasses were pinned down by the head on a silicone dish and immersed in hemolymph-like 3 (HL3) saline (Stewart et al., 1994). Thoraces were sagittally bisected, washed thrice in HL3 saline, and mounted with a 0.12 mm spacer. Imaging was performed immediately using a HC PL APO 63×/1.40 OIL CS2 objective on a Leica SP8 point-scanning confocal microscope. Five pre-bleach frames (1.29 seconds each) were recorded before bleaching with a 488-nm laser to achieve a bleach depth averaging 76%±7% (SD). Seventy-five post-bleach frames (1.29 seconds each) were then acquired. ROIs (1.25×3 μm) were placed over individual Z-discs and those with drift were discarded. For each fly, three Z-discs on different myofibrils were bleached, and intensities of adjacent unbleached Z-discs and unbleached background regions were recorded. Fluorescence was normalized as (bleached–background)/(unbleached– background), averaged across the three ROIs, and scaled such that the post-bleach frame equaled 0 and the pre-bleach mean equaled 100. Recovery curves were presented as mean±SEM. Each fly’s curve was fitted to a one-phase exponential association model to obtain the plateau (mobile fraction). Fixed samples were prepared in the same manner except with an additional ten-minute fixation in 4% paraformaldehyde in HL3 saline.

Microscopy

Confocal images were acquired using a HC PL APO 63×/1.40 OIL CS2 objective on a Leica SP8 point-scanning confocal microscope. The position along the z-axis was adjusted until perfectly in the centre of the myofibril, i.e., the widest point. Polarized light images were acquired using a 5X N-ACHROPLAN, NA=0.13 objective on a Zeiss Axio Observer microscope. TEM images were acquired using an FEI Tecnai G2 Spirit Twin 120 kV Cryo-TEM using an AMT NanoSprint15 MK2 CMOS camera.

Computational analyses

Schematics were illustrated with IBS 2.0 (Xie et al., 2022). All protein domains were defined per SMART (Letunic & Bork, 2018). Graphing and statistical analysis was performed using GraphPad Prism 10. Statistical analyses were performed in GraphPad Prism 10, where ns=p≥0.05, *=p<0.05; **=p<0.01, and ****=p<0.0001, and calculated to four decimal places. Equivalence of bend angles was assessed using the Two One-Sided Test (TOST) procedure. Model sarcomeres with bend angles ranging from 180° (straight) to 95° and blurring were generated. Each was hand measured ten times and the average absolute measurement error (4.5°) was used as the equivalence margin for all TOST analyses.

Minus strand RNA-seq bigWig files from the ENCODE project (Duff et al., 2015) with the following accessions were retrieved and loaded in IGV 2.19.4 (Robinson et al., 2011): ENCFF289ZZT, ENCFF982OGI, ENCFF430HWM, ENCFF716DNU, ENCFF829OGQ, ENCFF936YJM, ENCFF987YVI, ENCFF033MIY, ENCFF845HII, ENCFF021VXY, ENCFF792FNT. Signal tracks were aligned to the dm3 assembly and the data range was adjusted uniformly for all tracks.

SRA files with the following accession numbers were downloaded from the Bioproject PRJNA419412 (Spletter et al., 2018): SRR6314253, SRR6314256, SRR6314275, SRR6314259, SRR6314277, SRR6314262, SRR6314273. Salmon 1.10.0 (Patro et al., 2017) was used to quantify expression of the following transcripts which correspond to identified Zasp52 isoforms: FBtr0392902, FBtr0329914, FBtr0329912, FBtr0100387, FBtr0111048, FBtr0087315, FBtr0329913, FBtr0310086, FBtr0329916, FBtr0329915, FBtr0310088, FBtr0310087, FBtr0392901, FBtr0329911, FBtr0342711, FBtr0100388, FBtr0302163, FBtr0301313, FBtr0479884, FBtr0329909, FBtr0392918, FBtr0329910. The first six were classified as containing only exon 15a, the next six containing exons 15b-d, the next four containing exon 15e, and the last six containing no exon 15.

ImageJ was used to evaluate bending manually using the angle tool, and sarcomere length using the line measuring tool. For measuring patchiness, Niblack auto local thresholding of the actin channel using a radius of 15 units was performed, and outliers removed. The area of background within the myofibrils was divided by the area of the foreground thresholded selection to compute a patchiness metric.

For disorder prediction, the entire Zasp52-PF sequence was entered into AIUPred v0.1, DISOPRED3 (as two sequences, the first covering the first 1200 amino acids and the second, the remaining), or metapredict online v3.0. Raw values for each residue were obtained and smoothed using a Savitzky-Golay filter with fourth order smoothing and 11 neighbours on each side. AlphaFold v2.0 was used to obtain pLDDT scores for the alpha carbon of each residue. pLDDT scores were min-max normalized and smoothed in the same manner as above.

Zasp52 orthologs were obtained for members of the class Insecta using NCBI’s Eukaryotic Genome Annotation Pipeline. Canonical protein sequences for each were downloaded and those which did not contain the classical PDZ and four LIM domains were removed. The number of residues between the first and second LIM domains was calculated. A phylogenetic tree of these species was created using NCBI’s Common Tree (Schoch et al., 2020) and a heatmap created using iTOL v.6 (Letunic & Bork, 2024).

NARDINI (Cohan et al., 2022) was used for the sequences indicated to generate a z-score matrix, with 100,000 scrambled sequence variants.

eLife Assessment

This study investigates the role of the Z-disc protein Zasp52 in Drosophila flight muscles and provides evidence that an intrinsically disordered region (IDR) helps to stabilize and promote the localization of the protein to the Z-disc. Overall, this represents an important study that provides insights into Z-disc function and maintenance. The data are convincing, supported by strong genetic evidence and behavioral tests, well-controlled experiments, and detailed statistical analyses. Additional functional analyses designed to tease out specialized regions within the newly described isoform of Zasp52 would further strengthen models regarding the function of the protein.

Reviewer #1 (Public review):

The manuscript by Ho and Schock investigates the role of the Z-disc protein Zasp52 during Drosophila flight muscle development. It was known before, mainly by findings from this group, that Zasp52 is required for normal sarcomere morphogenesis, specifically Z-disc morphogenesis in indirect flight muscles. But the exact molecular mechanism by which Zasp52 contributes, apart from the fact that it is localised there and is somehow involved in multimerization/cross-linking, was not clear. This paper proposes that an intrinsically disordered region (IDR) in Zasp52 is needed for some of its functions, by stabilising Zasp52 localisation at the Z-disc. Specifically, the IDR in Zasp52 is proposed to be required for Z-disc maintenance during the mechanical challenges of flight, while being dispensable for the initial morphogenesis during development. This hypothesis is supported by strong genetic evidence and behavioural tests, deleting Zasp's IDR impairs flight from mid-age onwards, while a block in flight activity lifts the phenotype.

However, some of the phenotypic analysis, in particular the bending of the sarcomere, likely upon mechanical challenge by muscle contractions, needs more detailed investigations to be fully convincing.

Strengths:

(1) The linker in the alternatively spliced exon 15 of Zasp52 was deleted with a state-of-the-art genetic editing strategy. Surprisingly, flies are homozygous viable, showing that this long part of the Zasp52 protein is not essential for animal survival or sarcomere morphogenesis.

(2) The observed sarcomere phenotypes with age, especially the bending Z-discs, are new and exciting.

(3) The displayed EM images document interesting phenotypes.

(4) Most of the observed phenotypes can be rescued by re-expression of the long Zasp52 isoform, which does contain the IDR region, but not by a shorter one without it, suggesting that IDR is important.

(5) FRAP data measure the local turnover of a short-ZaspGFP and show that this increased in the Zasp mutant lacking the IDR domain, suggesting that Zasp-IDR might stabilise Zasp at the Z-disc.

(6) Interestingly, flight and sarcomere morphology phenotypes can be rescued by preventing the flies from flying, suggesting that they are mechanically induced.

Weaknesses:

(1) The western blot quantifications of Zasp isoform expression are weak. No error bars are indicated in the quantifications; the quantifications appear to be more qualitative than quantitative. According to band intensities, the long Zasp isoforms seem to be less present compared to the shorter ones, even in the flight muscles.

(2) The phenotypic analysis of the sarcomere appears somewhat superficial throughout the paper. Only Zasp52 and phalloidin are shown; no other Z-disc or thick filament proteins. At least myosin stainings and overview images are important to better judge the phenotypic variations. Are the variants between individuals or regional in the same muscle?

(3) EM images would benefit from better quantification.

(4) Other proteins were not analysed with the FRAP-based turnover assay for comparison in wild type and mutant. All Z-proteins might turn over faster in the mutant with the defective Z-disc.

Reviewer #2 (Public review):

Summary and Strengths:

This in-depth genetic analysis of Zasp52 function in Drosophila indirect flight muscle (IFM) provides an interesting perspective regarding the role of a partially disordered region (IDR) in exon 15e. This exon seems to be exclusively present in IFM and contributes to the prevention of myofibril disintegration during aging, likely due to interactions of this region with Z-disc insertion and/or stability. The addition of an isoform (PR) that lacks exon 15e serves as a nice control to illustrate the necessity of exon 15e in muscle structure and function. Overall, the manuscript is exceptionally well-written, logical, with nicely controlled experiments and detailed statistical analysis that largely support the conclusions drawn by the authors. While exon 15e is clearly involved in preventing muscle degeneration, a solid role for thin filament stability is not clearly shown (as mentioned in the abstract). In addition, which regions/how the proteins of the IDR may contribute are unclear.

Weaknesses:

(1) It is not clear in Figure S1A where exon 15e fits within the Zasp52 locus schematic. This is important as a premise of this paper describes this region to be key, and proof from multiple prediction programs would lend more weight to the prediction of the exon being largely disordered. Inclusion of the discussed short linear motifs, comparison with Canoe or LBD3 for similarities and/or an Alphafold structure would help make the authors' point (colorized with known domains).

(2) Interesting that immobilization rescues the deterioration phenotypes. The authors should explain in more detail how this was done to avoid dehydration/starvation of the flies.

(3) There is a lot of discussion about the potential function of the IDR region, specifically a putative actin binding motif or other 'ordered' regions that may contain short linear motifs. It would strengthen the findings to show which of these may be essential for Zasp52 function in the IFM. The ability to bind actin could be tested biochemically, and/or smaller deletions could be made to unequivocally test the role of the ABD vs other predicted motifs using genetics. If some of these regions are more ordered, where do they lie within, and do they form a predicted fold or structure that gives insight into function?

Author response:

The following is the authors’ response to the original reviews.

Thank you for the helpful comments and criticisms. We provide exciting additional data, in particular a CRISPR actin-binding motif mutant and FRAP analysis of an exon15e-GFP transgene, both further supporting the importance of the IDR in thin filament stability. We believe that these additional experiments provide compelling evidence supporting our conclusion and substantially advance the current limited body of knowledge surrounding the role of IDRs in structural proteins.

Public Reviews:

Reviewer #1 (Public review):

The manuscript by Ho and Schock investigates the role of the Z-disc protein Zasp52 during Drosophila flight muscle development. It was known before, mainly by findings from this group, that Zasp52 is required for normal sarcomere morphogenesis, specifically Z-disc morphogenesis in indirect flight muscles. But the exact molecular mechanism by which Zasp52 contributes, apart from the fact that it is localised there and is somehow involved in multimerization/cross-linking, was not clear. This paper proposes that an intrinsically disordered region (IDR) in Zasp52 is needed for some of its functions, by stabilising Zasp52 localisation at the Z-disc. Specifically, the IDR in Zasp52 is proposed to be required for Z-disc maintenance during the mechanical challenges of flight, while being dispensable for the initial morphogenesis during development. This hypothesis is supported by strong genetic evidence and behavioural tests, deleting Zasp's IDR impairs flight from mid-age onwards, while a block in flight activity lifts the phenotype.

However, some of the phenotypic analysis, in particular the bending of the sarcomere, likely upon mechanical challenge by muscle contractions, needs more detailed investigations to be fully convincing.

Strengths:

(1) The linker in the alternatively spliced exon 15 of Zasp52 was deleted with a state-of-the-art genetic editing strategy. Surprisingly, flies are homozygous viable, showing that this long part of the Zasp52 protein is not essential for animal survival or sarcomere morphogenesis.

(2) The observed sarcomere phenotypes with age, especially the bending Z-discs, are new and exciting.

(3) The displayed EM images document interesting phenotypes.

(4) Most of the observed phenotypes can be rescued by re-expression of the long Zasp52 isoform, which does contain the IDR region, but not by a shorter one without it, suggesting that IDR is important.

(5) FRAP data measure the local turnover of a short-ZaspGFP and show that this increased in the Zasp mutant lacking the IDR domain, suggesting that Zasp-IDR might stabilise Zasp at the Z-disc.

(6) Interestingly, flight and sarcomere morphology phenotypes can be rescued by preventing the flies from flying, suggesting that they are mechanically induced.

Weaknesses:

(1) The western blot quantifications of Zasp isoform expression are weak. No error bars are indicated in the quantifications; the quantifications appear to be more qualitative than quantitative. According to band intensities, the long Zasp isoforms seem to be less present compared to the shorter ones, even in the flight muscles.

We have now included quantifications with error bars for the Western blots in our resubmission. It is important to keep in mind that the main point in figure 1B is that there are plenty of exon15e-containing isoforms in IFM, in contrast to other tissues with very limited exon15e-containing isoforms. This is confirmed by the analysis of RNA-seq data in figure 1C, and of course, by the flightless phenotype of the exon15e mutant.

(2) The phenotypic analysis of the sarcomere appears somewhat superficial throughout the paper. Only Zasp52 and phalloidin are shown; no other Z-disc or thick filament proteins. At least myosin stainings and overview images are important to better judge the phenotypic variations. Are the variants between individuals or regional in the same muscle?

Our images are representative of the observed phenotypes. Phenotypes are consistently present across all individuals, as reflected in our replicates. Interestingly, they appear not to be randomly interspersed among the sarcomeres but concentrated in certain regions of muscle more than others. Full images are available in the online repository FigShare.

(3) EM images would benefit from better quantification.

We do not believe that EM images can be meaningfully quantified, because of the many selection steps preceding image acquisition.

(4) Other proteins were not analysed with the FRAP-based turnover assay for comparison in wild type and mutant. All Z-proteins might turn over faster in the mutant with the defective Z-disc.

This is the point we are trying to make. The Zasp52 IDR appears to stabilize the Z-disc and is likely involved in fastening a variety of proteins to it.

Reviewer #2 (Public review):

Summary and Strengths:

This in-depth genetic analysis of Zasp52 function in Drosophila indirect flight muscle (IFM) provides an interesting perspective regarding the role of a partially disordered region (IDR) in exon 15e. This exon seems to be exclusively present in IFM and contributes to the prevention of myofibril disintegration during aging, likely due to interactions of this region with Z-disc insertion and/or stability. The addition of an isoform (PR) that lacks exon 15e serves as a nice control to illustrate the necessity of exon 15e in muscle structure and function. Overall, the manuscript is exceptionally well-written, logical, with nicely controlled experiments and detailed statistical analysis that largely support the conclusions drawn by the authors. While exon 15e is clearly involved in preventing muscle degeneration, a solid role for thin filament stability is not clearly shown (as mentioned in the abstract). In addition, which regions/how the proteins of the IDR may contribute are unclear.

Weaknesses:

(1) It is not clear in Figure S1A where exon 15e fits within the Zasp52 locus schematic. This is important as a premise of this paper describes this region to be key, and proof from multiple prediction programs would lend more weight to the prediction of the exon being largely disordered. Inclusion of the discussed short linear motifs, comparison with Canoe or LBD3 for similarities and/or an Alphafold structure would help make the authors' point (colorized with known domains).

We added a bar below figure S2A to show the region corresponding to exon 15e. We used three disorder prediction programs and one structure (order) prediction program. The majority of exon15e is completely disordered and of very low confidence score, and thus uninformative to display as an AlphaFold structure. Likewise, IDR’s are very difficult to classify, therefore we cannot say much more than that LDB3, Zasp52, and Canoe contain IDRs, with Zasp52 and Canoe both having a putative actin-binding domain within the IDR. We now provide data on the function of the ABD in this resubmission.

(2) Interesting that immobilization rescues the deterioration phenotypes. The authors should explain in more detail how this was done to avoid dehydration/starvation of the flies.

We provided more details in materials and methods.

(3) There is a lot of discussion about the potential function of the IDR region, specifically a putative actin binding motif or other 'ordered' regions that may contain short linear motifs. It would strengthen the findings to show which of these may be essential for Zasp52 function in the IFM. The ability to bind actin could be tested biochemically, and/or smaller deletions could be made to unequivocally test the role of the ABD vs other predicted motifs using genetics. If some of these regions are more ordered, where do they lie within, and do they form a predicted fold or structure that gives insight into function?

We now provide data on the function of the ABD showing that deleting it has almost no phenotypic defects. That means the IDR is largely/entirely responsible for the observed phenotypes.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Western blot in Figure 1B needs proper quantification. A ratio between long and short isoforms in the same muscle type might be informative. Is it known which epitope the antibody recognises? Can a GFP insertion that also labels all isoforms be used as verification? Quantifications are also needed in Figure 2A.

We have added quantifications of all Western blots (Fig. 1B, 2A, and 2A’). The ratio between exon 15e-containing and total Zasp52 in the same muscle type is included in the lowest bar graph in Fig. 1B. The full-length antibody is polyclonal and was raised against Zasp52-PR which contains all ordered domains; the anti-LIM antibody was raised against the last three LIM domains (both are described or referenced in the materials and methods section). Such a GFP insertion cannot exist due to the complex splicing patterns of Zasp52.

(2) The name of the deletion allele could be specifically indicated in Figure 1A below the red bar.

Done.

(3) It would be useful to indicate the order group names in Figure S2B since species names are hard to read.

For the version of record we provided high-resolution images, where species names can be read. Drosophilids, Ephemeroptera and Odonata are indicated.

(4) The inverted spelling of the numbers for the control in Figures 2C and 5H is strange.

Changed to normal spelling.

(5) The bending of the myofibril at the Z-disc is a really interesting phenotype. However, it seems it is not always visible; at least it is visible in many myofibrils shown in Figure 3B, but in none in Figure 3E, same genotype, just different staining. Hence, I wonder if this bending could be force-induced by the cutting of the thorax during tissue preparation. It would be useful to display some overview images to allow the reader to judge the quality of the tissue preparation, indicating from where the high magnification view shown was taken. The same is true for Figure 5.

Overview images are available on FigShare. Note that you can see some “H-zone actin” sarcomeres in Fig. 3B, as well as some mildly bent ones in Fig. 3E. We generally selected images that best demonstrated the phenotype described. Furthermore, neither phenotype is fully penetrant so we cannot expect to see it everywhere. Lastly, it is always possible that phenotypes are affected by preparation, since it is impossible to know what the myofibrils look like in situ. However, all samples were prepared using the same protocol with replicates, and since we see a phenotype in our mutants and not in the control, this indicates that something is different between the two.

(6) The same applies to the visualisation of the "hyper-contracted" phenotype; again, it seems to be an all-or-nothing phenotype in the zoom shown. An overview image should be shown. The zoom in Figure 4E would benefit from displaying phalloidin in a separate channel. Are actin filaments pulled out of the Z-disc? The latter is often seen in non-perfect cuts in wild-type, but the accumulation at the M is curious. It would be informative to locate the ends of the thick filaments in these cases or quantify thick filament lengths; do these invade the Z-discs? This can easily be done by a myosin staining.

Is this a regional effect or does it depend on the individual or on the preparation? I am surprised to also see the "hyper-contraction" in 10% of wild-type 5-day adults.

See previous response where we include overview images. Single-channel images are available; it is visible that actin filaments are not pulled out of the Z-disc. Phenotypes are consistent across individuals as evidenced in our replicates but do tend to be concentrated in certain regions of muscle.

(7) The EM images would benefit from more overview images. At the moment, we only see a single sarcomere from wild type and mutant, with no quantification of the phenotype. Can the authors see the invading thin filaments into the M-band? The disrupted Z-disc phenotypes are impressive. What is the age of the animal shown in Figure 4?

We have a panel displaying several mutant sarcomeres. Due to the selectivity and challenges of the EM preparation process, we do not believe we can perform meaningful statistics on them. It sometimes looks like myosin heads are visible in the H zone which may support the presence of thin filaments in the H zone (Fig. 4B and C). However, the quality of these particular EM images is not high enough to identify thin filaments. All phenotypes shown are from 3-week-old animals.

(8) Is UH-3 GAL4 expressed at the adult stage?

Yes, from 36 h APF into adulthood (Singh et al. 2014). Now mentioned in the results section.

(9) Figure 6 would strongly benefit from a myosin staining. Do thin and thick filament lengths scale? It seems that overlap is reduced in the double hets. How can this be envisioned with Z-disc stability? Is myofibril diameter reduced?

We searched for non-additive differences in myofibril diameter but were unable to detect any.

(10) What is the FRAP turnover rate of a long Zasp-GFP compared to a short one in wild type? A difference would indicate that it is really the IDR domain that keeps Zasp52 longer at the Z-disc, instead of an indirect effect caused by Z-disc morphology

We have newly added FRAP data of a GFP-tagged exon 15e construct which displays much lower turnover. This indicates that the IDR does indeed retain Zasp52 at the Z-disc.

Reviewer #2 (Recommendations for the authors):

(1) The total protein stain should also be included if it is used for quantitation in Figures 1B and 2A-A'.

These are available on FigShare.

(2) It is a bit confusing that the Alphafold plot is inversely correlated with the other 3 prediction programs, although this is explained in the legend. Maybe an Alphafold structure would help make the authors' point (colorized with known domains).

The AlphaFold structure is almost entirely low-confidence disordered region except for the structured domains so we do not believe it would be helpful to include.

(3) The title of Figure 8 says 'Certain ex15e defects are rescued by immobilization.' What other defects are not rescued? If true, these should be shown.

There was a full rescue. We deleted the word “certain”

(4) Please include a brief explanation of the spatiotemporal expression of UH3-Gal4.

From 36h APF into adulthood (Singh et al. 2014). Now mentioned in the results section.

(5) Statistics should be added to Figure 8E.

Figure 8E (now 9E) has statistics.

(6) The dark blue color used for integrin staining in Figure S3 is difficult to see. Changing this color may help visualize differences. Also, pointing them out with arrows, etc., will help clarify abnormalities.

We have described these differences in the figure caption. Single-channel images are available for viewing in any color in FigShare.

Version 2

Abstract

The Drosophila scaffolding protein Zasp52 is required to maintain structure at the muscle Z-disc which experiences strong forces during contraction. It is alternatively spliced into many isoforms, some of which contain a long intrinsically disordered region (IDR). We show that this region is primarily expressed in the indirect flight muscle (IFM) and is required for maintaining integrity of the Z-disc. Deleting the IDR-encoding exon 15e results in flightlessness and structural IFM defects, including sarcomere bending at the Z-disc and an inability to de-contract. These defects are indicative of a lack of proper thin filament anchoring to the Z-disc. This is further supported by a genetic interaction between exon 15e and actin. Fluorescence recovery after photobleaching of an isoform lacking exon 15e shows that the IDR is required for maintaining Zasp52 at the Z-disc and thereby stabilizing Z-discs. Lastly, we can rescue these phenotypes by restricting IFM use. Together, these results suggest that Zasp52’s IDR confers thin filament stability at the Z-disc of IFM.

Background

The fruit fly Drosophila melanogaster possesses 152 muscle groups in the imago, as well as 54 in the larva (Bothe & Baylies, 2016). Each muscle is adapted to perform a specialized function, such as the extremely ordered indirect flight muscle (IFM) which is designed so as to enable the rapid oscillatory contractions needed to sustain flight. The marked phenotypic differences among these muscle types begs a molecular explanation. Alternative splicing is known to contribute to muscle type development and function with about 80% of sarcomere proteins being differentially expressed or spliced in Drosophila muscle (Nikonova et al., 2020). One such protein is Zasp52, a scaffolder, which contains numerous alternatively spliced exons generating a variety of isoforms.

All metazoan muscles share considerable structural homology (Steinmetz et al., 2012). They come in two types: striated, which appear striped given the presence of sarcomeres, and smooth, which do not possess sarcomeres. The sarcomere, a cylindrical structure, is the fundamental contractile unit of striated muscle and is bordered on either side by a dense circular structure called the Z-disc. Sarcomeres are arranged end-to-end in long arrays called myofibrils. Thin filaments composed of actin are anchored at and extend perpendicularly from the Z-disc into neighbouring sarcomeres, terminating at the H-zone (the region halfway between two Z-discs with no thin filaments). Thick filaments composed of muscle myosin are centred at the M-line (found at the centre of the H-zone) and interdigitate with thin filaments in a hexagonal lattice. The mechanism of sarcomere contraction is explained by the sliding filament theory in which thick filaments pull upon thin filaments in an ATP-dependent manner, bringing the Z-discs closer together. The simultaneous contraction of tandem sarcomeres comprising a myofibril is what leads to large-scale muscular contraction (Krans, 2010). The Z-disc is crucial for maintaining mechanical stability of the muscle and contains numerous proteins including Zasp52 which interacts with the thin filament cross-linker alpha-actinin (Liao et al., 2016) and actin (Liao et al., 2020) in Drosophila.

Zasp52, part of the three-membered Drosophila Alp/Enigma protein family along with Zasp66 and Zasp67, is crucial for maintaining structure at the Z-disc. It localizes to the Z-disc in all muscle types of Drosophila melanogaster (Katzemich et al., 2011) and is required for the assembly and maintenance of IFM sarcomeres (Katzemich et al., 2013; Liao et al., 2016; Medina-Quintana et al., 2026). Developmentally, Zasp52 appears in the embryo first at myotendinous junctions and also at Z-bodies - clusters which later coalesce to form Z-discs (Katzemich et al., 2013). Zasp52 binds actin via its N-terminal PDZ domain plus a long C-terminal extension (Liao et al., 2020), and α-actinin via a region encompassing the PDZ domain plus a short C-terminal extension (Liao et al., 2016). In addition to the PDZ and Zasp motif (ZM) domain, Zasp52 contains four LIM domains which interact with the ZM domain resulting in auto-oligomerization; this mechanism is believed to drive radial growth of the Z-disc as well as pathological aggregate formation (González-Morales et al., 2019; Medina-Quintana et al., 2026). Its primary human ortholog LDB3/ZASP is implicated in muscular dystrophies which are characterized by progressive adult-onset distal myofibrillar myopathy and sometimes cardiomyopathy (Griggs, et al., 2007).

Zasp52 is alternatively spliced into twenty-two different isoforms (according to FlyBase release FB2025_02, used throughout this paper), which leads to a variety of different domain and inter-domain combinations. The inter-domain linker regions connecting Zasp52’s structured domains (i.e., PDZ, ZM, and LIM domains) are less characterized but are proposed to play an important role in protein function (Fisher & Schöck, 2022).

Though not much is known about the linker region between LIM1 and LIM2, its predominant expression in IFM and unusually large length prompted us to investigate it. We were curious to see how deleting only this linker region while retaining all conserved domains would affect muscle structure and function. This linker was found to be highly disordered; intrinsically disordered regions (IDRs) have mostly been analyzed biochemically, with only limited data on their in vivo function in model organisms (Jensen et al., 2025). Flight ability was impaired and myofibrils displayed a variety of phenotypes which were all indicative of a loss of thin filament anchoring at the Z-disc. Most phenotypes exhibited an age-related increase in severity. Put together, these data demonstrate a crucial role for this large IDR in maintaining sarcomere integrity by stabilizing thin filaments at the Z-disc in IFM.

Results

exon 15e-containing isoforms are expressed in a specific spatiotemporal manner

Zasp52 isoforms containing the first and second LIM domains contain a linker region between the two which can vary greatly in size. Encoded by several exons, this linker region may contain exon 15 which undergoes alternative 5’ splice site selection generating five exonic variants (15a-e) ranging from 130 nucleotides (15a) to 4429 nucleotides (15e) in length (Figure 1A). The last 29 nucleotides of exon 15 contribute to the second LIM domain, but the remainder encodes the linker. When exon 15e is included, it constitutes the majority of the entire protein, encoding 1475 amino acids which total 160 kDa. In the canonical and longest isoform Zasp52-PF, the entire protein is 2194 amino acids long and the linker region between LIM1 and LIM2 is 1686 amino acids long. Exon 15e is referred to variously by other sources: as exon 9 in Flybase, exon 16c in Katzemich et al., 2011, and exon 18/19 in Benna et al., 2009; protein isoforms containing exon 15e were previously also referred to as Z(210) (Chechenova et al., 2013).

exon 15e-containing isoforms are expressed in a specific spatiotemporal manner.

(A) A schematic of the Zasp52 locus. Exons are drawn to scale, introns are not. Coding exons are shown in orange and untranslated regions in grey. Alternative start sites are also indicated. Below in green are mappings of structured domains according to SMART (Letunic & Bork, 2018); and also Katzemich et al, 2011 which defined LIM1 and LIM2 variants. The deletion in ex15e mutants is indicated in red. (B) Western blot using five different tissue extracts from control flies probed with a Zasp52 full-length antibody. Relative amounts of Zasp52 content as measured in each lane, over the total protein content as measured using the TGX Stain-free gel method, are quantified below for each tissue as mean with error bars representing standard deviation (n=3). Exon 15e content as measured using the protein content above the dotted blue line for each lane, over Zasp52 content is quantified below that. Molecular weights appear higher, perhaps due to retardation caused by non-specific interactions with the gel matrix, as also seen in Watts et al., 2017, Figure S1. (C) RNA-seq data from Spletter et al., 2018 of IFM tissue across various timepoints after puparium formation; the last timepoint is of adults one day after eclosion. Gene expression levels in transcripts per million (TPM) are shown for Zasp52 isoforms classified according to their splicing of exon 15. The majority of Zasp52 IFM isoforms contain exon 15e.

Western blotting using a polyclonal antibody capable of detecting all Zasp52 splice variants (hereinafter referred to as the Zasp52 full-length antibody, Jani & Schöck, 2007) was used to analyze the presence of Zasp52 in IFM, ovaries, whole larvae, heads, and whole male abdomens (Figure 1B). IFM contains the second most Zasp52 protein of any isoform relative to total protein content compared to the other tissues analyzed and is the only tissue that significantly contains exon 15e-containing isoforms relative to total Zasp52 content. This agrees with previous studies (Chechenova et al., 2013, Benna et al., 2009) which found no higher weight isoforms in larvae, and with transcriptomics data. Whole organism RNA-seq data from the ENCODE project (Duff et al., 2015) reveals very little expression of exon 15e in all developmental stages up until four days past the white prepupal stage (Figure 1 – Figure supplement 1). To verify that this is indeed a phenomenon of exon 15e inclusion and not merely Zasp52 expression, one can look at the surrounding exons which do not follow the same pattern as exon 15e. To gain a more precise perspective on IFM expression of exon 15e, we turned to an RNA-seq dataset from Spletter et al., 2018 which looked at total RNA content of IFM at various developmental time points (Figure 1C). Expression of all Zasp52 transcripts in IFM surges from 48 to 72 h after puparium formation (APF), with most of these transcripts containing exon 15e, as opposed to including exon 15 a-d only or no exon 15 at all.

We used various software to predict the level of disorder across Zasp52-PF (Figure 1 – Figure supplement 2A). All identified the linker region between LIM1 and LIM2 as an intrinsically disordered region (IDR) to various degrees, although some regions within it appear more ordered. The actin-binding motif (ABM) identified by Ashour et al., 2023 within exon 15e is predicted to be more structured than the surrounding regions. No significant sequence similarity was found when aligning Zasp52’s linker region to human or other Drosophila melanogaster ALP/Enigma protein family members. We then compared the lengths of the linker regions between LIM1 and LIM2 in all insect species possessing a Zasp52 ortholog as defined by NCBI’s Eukaryotic Genome Annotation Pipeline (Figure 1 – Figure supplement 2B). Drosophilids generally have a longer linker than their taxonomical neighbours. Mayflies, dragonflies, and damselflies, which use direct flight muscle, have notably short linkers. Since linear sequence often does not fully describe an IDR’s function, we analyzed linear clustering or mixing of different residue types using NARDINI (Cohan et al., 2022). The z-score matrix of the linker region between LIM1 and LIM2 in Zasp52-PF showed a strong clustering/segregation of polar residues (µ), as well as alanine (A) and proline (P), and hydrophobic residues (h) (Figure 1 – Figure supplement 2C). The z-score matrix of exon 15e alone displayed much of the same clustering, except with additional strong clustering of acidic residues (-) with respect to other residue classes (Figure 1 – Figure supplement 2C’). This implies that the additional exons in the linker region dilute or mask the strong clustering of acidic residues in exon 15e alone. We attempted to compare z-score matrices across various arthropod species but were unable to detect any meaningful correlation indicating that the binary patterns uncovered by NARDINI do not reveal taxonomical differences in exon 15e composition.

A CRISPR deletion of exon 15e displays flight defects

To further investigate the function of exon 15e, we generated a CRISPR deletion spanning five intronic nucleotides five prime to exon 15e up to 149 nucleotides before the 3’ end of exon 15e (Figure 1A). In total, 4285 nucleotides totaling 156 kDa were deleted. Thus, our mutants would affect all isoforms containing exons 15b-e. The deletion endpoints were sequenced to confirm effective CRISPR knockout. To confirm that our deletion worked and spliced as intended, we performed Western blots using the Zasp52 full-length antibody to compare our deletion with a control and the MiMIC line MI00979 (Venken et al., 2011) which truncates everything C-terminal to and including exon 15e (Figure 2A). The largest protein that could possibly be present in our mutant would be 136 kDa (Zasp52-PI). Our deletion was devoid of all higher molecular weight isoforms containing exon 15e which were present in the control as expected. Our deletion also contained several isoforms which were heavier than the highest molecular weight isoforms of MI00979 which were also not present in control, suggesting that these are novel products containing the C-terminal LIM domains. To confirm that this was indeed the case, i.e., that the three C-terminal LIM domains were retained, we raised an antibody against them and performed a Western blot (Figure 2A’). Our CRISPR deletion retained these LIM domains as expected. We refer to this CRISPR mutant as ex15e.

A CRISPR deletion of exon 15e displays flight defects.

Western blots using either the Zasp52 full-length antibody (A) or antibody against LIM2-LIM4 (A’) of either w1118 (ctrl hereinafter), ex15e, or the MiMIC line MI00979 whole flies. Relative amounts of Zasp52 content as measured in each lane, over the total protein content, are quantified below for each genotype as mean with error bars representing standard deviation (n=3) normalized against the ctrl lane. (B) Larval crawling assay of either ctrl (n=21) or ex15e (n=21) third instar larvae showing no significant difference between the two (unpaired t-test with Welch’s correction; p=0.6006). (C) Flight assays of ctrl vs ex15e flies of three different age categories: one-day-old (ctrl n=138 flies, ex15e n=123), five-days-old (ctrl n=125, ex15e n=119), and three-weeks-old (ctrl n=86, ex15e n=81). The y-axis indicates flight strength: flies were released into a tube and those that landed in the top segment (y=1) had the strongest flight strength while those that landed in the bottom dish (y=8) had the weakest. The x-axis indicates the proportion of flies that landed in that segment. Red arrows indicate the average position landed in ctrl flies, yellow arrows are for ex15e flies. Difference in flight ability was statistically significant for all age categories (Fisher’s exact test; p<0.0001).

We wanted to see how ex15e affected the locomotory abilities of the fly. Since larvae do not express exon 15e, we do not expect them to be affected by our deletion. Therefore, we performed a larval crawling assay as a negative control to confirm that ex15e larval muscles were unaffected (Figure 2B). As predicted, control and ex15e larvae demonstrated no differences in crawling ability. Next, we wanted to see how flight would be affected, given that exon 15e is present primarily in the adult IFM. ex15e flies of all age categories had impaired flight ability compared to control with a severe deterioration at three weeks of age (Figure 2C). To ensure that no secondary-site mutations were affecting our mutant phenotypes, we performed a simple binary flight assay using the heterozygous ex15e deletion over a Zasp52 deficiency (Figure 2 – Figure supplement 1A).

This impairment in flight could stem from a variety of causes, given the multimodal functionality of Zasp52. We decided to first look at gross muscle morphology before zooming in on our investigation. Thorax bisections revealed no noticeable difference in the IFMs of ex15e and control flies (Figure 2 – Figure supplement 1B, C). Since Zasp52 was known to be involved in integrin interactions (Jani & Schöck, 2007; Bouaouina et al., 2012) and junctional components (Ashour et al., 2023), we inspected the myotendinous junctions of IFM, which are called modified terminal Z-discs (MTZ), in three-week-old flies (Figure 2 – Figure supplement 1D-G’’). βPS integrin localization appeared unaffected in ex15e; this agrees with integrins being upstream of Zasp52 (Jani & Schöck, 2007). However, the MTZ appeared disrupted and malformed, lacking the organized structure seen in controls. Furthermore, we also noticed myofibril defects which prompted us to further investigate IFM at the myofibrillar level.

ex15e mutant IFM phenotypes: bending at the Z-disc and thin filament intrusion into the H-zone

IFM tissue was analyzed by confocal microscopy with several phenotypes being observed. First, myofibrils in ex15e mutants were often found to be bent at the Z-disc (Figure 3A-C). In control flies, sarcomeres across different age groups typically displayed the same proportion of degree of bentness. In ex15e, the proportion of bent sarcomeres increased with age, as well as the degree of bending, compared to control flies. Force-induced bending results in a sarcomere whose inner curve is bowed, much like a hose with a kink. The bending in ex15e appears to be a result of the improper anchoring of thin filaments which leads to a shear type angle shift in the sarcomere at the Z-disc. Bending was found exclusively at the Z-disc and not the M-line in all muscle analyzed.

ex15e mutant IFM phenotypes: bending at the Z-disc and thin filament intrusion into the H-zone.

(A-C) Analysis of bent sarcomere phenotype. Actin is visualized by phalloidin in green. Ctrl myofibrils (A) appear straight while ex15e myofibrils (B) contain a range of bending (representative three-week-old myofibrils shown). Red arrowhead indicates a representative sarcomere with a bend of <145°, yellow arrowhead indicates 145-165°, and white arrowhead >165°. (C) Bend angles were statistically equivalent (≡) between ctrl and ex15e in one-day-olds and five-day-olds, but not equivalent (ne) in three-week-olds, where ex15e displayed greater bentness. Sample sizes: one-day-old (ctrl n=1242 sarcomeres, ex15e n=1851), five-days-old (ctrl n=1192, ex15e n=946), and three-weeks-old (ctrl n=1768, ex15e n=1125) (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; 1d.o. tL=12.95, tU=-12.07; 5d.o. tL=6.03, tU=-8.01; 3w.o. tL=-6.34, tU=-20.63). (D-F) Sarcomeres also exhibited actin in the H-zone phenotype (representative three-week-old myofibrils shown). Ctrl myofibrils (D) display normal H-zones without actin with M-lines/H-zones indicated by blue arrowheads. ex15e myofibrils (E) display actin accumulation at the H-zone; arrowheads indicate same as above. Actin is visualized by phalloidin in green and Z-discs are identified by the full-length Zasp52 antibody in red. (F) These sarcomeres with H-zone actin, which appear hypercontracted, are significantly overrepresented in ex15e compared to ctrl across all age categories. Sample sizes were as follows: one-day-old (ctrl n=3255 sarcomeres, ex15e n=4056), five-days-old (ctrl n=3221, ex15e n=2677), and three-weeks-old (ctrl n=5394, ex15e n=4720 (Fisher’s exact test; p<0.0001). (G) To ascertain that sarcomeres with actin in the H-zone were indeed contracted, the Z-to Z-disc length was measured in ex15e three-week-old flies of both sarcomeres with a bare H-zone (H-zone normal, n=89 sarcomeres) and those with actin in the H-zone (H-zone actin, n=117). H-zone actin sarcomeres were significantly smaller (unpaired t-test with Welch’s correction; p<0.0001).

Another phenotype observed in ex15e was the presence of a dense band of actin staining at the H-zone, where normally there would be no actin staining (Figure 3D-F). Control flies have very few sarcomeres with this phenotype at all ages; ex15e are similar to control at one day of age but begin to accumulate a large number of these abnormal sarcomeres with time. At three weeks of age, a large proportion of sarcomeres have actin at the H-zone. Usually, groups of myofibrils in the same area would exhibit this phenotype rather than these abnormal myofibrils being randomly dispersed among the normal ones. This dense band of actin suggested that thin filaments were overlapping at the M-line. This could be a result of hypercontraction, where thin filaments are brought so close together that they overlap. However, the protocol used for these preparations includes a relaxing solution containing ATP, which induces cross-bridge detachment resulting in the adoption of a relaxed conformation. To see if these abnormal sarcomeres with actin in the H-zone were indeed contracted, we measured their length and compared them to normal sarcomeres from the same sample in three-week-old flies (Figure 3G). Sarcomeres with actin in the H-zone had a smaller length than those without indicating that they were indeed contracted. Muscle decontraction relies on the elastic nature of the sarcomere to restore it to a relaxed conformation once myosin heads detach from the thin filaments. This requires stable Z-discs and myofilament attachments, the lack of which may be the cause of a permanent hypercontracted state in ex15e.

To further investigate these phenomena, we performed transmission electron microscopy (TEM) of myofibrils and noticed phenotypes that concurred with our immunofluorescence analysis. Wild type IFM sarcomeres of three-week-old flies are highly stereotypic with electron-dense, continuous Z-discs and regular H-zones (Figure 4A). In contrast, ex15e mutants show low-density, discontinuous Z-discs with many gaps as well as irregular H-zones (Figure 4B). We also observed bent Z-discs and split myofibrils, as well as fraying of filaments at the sarcomere periphery (Figure 4C). The latter two phenotypes could also be observed in transverse sections of myofibrils (Figure 4D). Altogether, these observations indicate a breakdown of Z-disc coherence and stability leading to additional downstream effects like myofibril splitting and fraying.

Transmission electron microscopy reveals defects in mutant sarcomere ultrastructure.

TEM images reveal similar phenotypes as immunofluorescent ones (three-week-old shown). Ctrl (A) sarcomeres are very structured and even. Longitudinal sections of ex15e (B, C) display broken Z-discs (red arrowhead), broken M-lines (yellow arrowhead), bending at the Z-discs (green arrowhead), splits in filaments (purple arrowhead), and fraying of peripheral filaments (blue arrowhead). Transverse sections of ex15e (D) also display some of these abnormalities (purple and blue arrowhead).

A putative actin-binding motif within exon 15e is not responsible for ex15e defects

A putative F-actin binding motif (ABM) identified by Ashour et al., 2023 is located within exon 15e and might be responsible for the thin filament-related phenotypes we witnessed in ex15e (see Figure 1- Figure supplement 2A for location). This short motif found in the CapZbeta protein of various metazoans is present in the Zasp52 proteins of the D. melanogaster subgroup (Ashour et al., 2023). To test the role of this ABM, we used CRISPR to delete five amino acids (LQREL) which include two highly conserved leucines and confirmed the in-frame deletion by sequencing – we refer to this mutant as ΔABM. Myofibrils appear normal (Figure 5A-C) and do not display the abnormal bending (Figure 5D) and contraction (Figure 5E) of ex15e. Flight ability distributions are slightly different from control flies although the average flight index is almost identical (Figure 5F). In conclusion, ΔABM does not phenocopy ex15e which indicates that exon 15e’s function is not reducible to that short motif. Instead, the IDR is essential for the phenotypes we observed.

A putative actin-binding motif within exon 15e is not responsible for ex15e defects.

ΔABM homozygotes along with all other flies in this figure were analyzed at three weeks of age. ΔABM myofibrils (C) resembled those of ctrl (A) and showed none of the phenotypes seen in ex15e mutants (B). Actin is visualized by phalloidin in green. (D) Bend angles were equivalent between ctrl and ΔABM flies (n=1648) (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; tL=21.88, tU=-7.93). (E) The proportion of sarcomeres that were contracted was not significantly different between ctrl and ΔABM flies (n=3773) (Fisher’s exact test; p=0.1274). (F) Flight ability was only slightly decreased in ΔABM flies (n=104 flies) compared to ctrl (n=86) but significantly improved compared to ex15e (n=81). Red arrows indicate the average position landed in ctrl flies, yellow are ex15e, and green are ΔABM (Fisher’s exact test; ctrl-ΔABM p=0.0351; ctrl-ex15e and ex15e-ΔABM p<0.0001).

ex15e defects are restored by rescue constructs

To determine if the phenotypes are truly due to the IDR or are caused by an altered isoform composition, we wanted to see if the phenotypes of ex15e could be rescued with two different isoforms: Zasp52-PF is the canonical full-length Zasp52 isoform which contains exon 15e, while Zasp52-PR contains all structured domains but not exon 15e (Figure 6A). We expressed either UAS-Zasp52-PF or UAS-Zasp52-PR using the UH3-GAL4 driver in a homozygous ex15e background and examined three-week-old flies (Figure 6B-H). UH3-GAL4 is expressed in IFM from 36 hours APF into adulthood (Singh et al., 2014). The degree of bending in Zasp52-PR rescue sarcomeres was statistically equivalent to that of ex15e mutants, while that of Zasp52-PF rescues was equivalent to control flies, indicating that Zasp52-PF but not Zasp52-PR can rescue this phenotype (Figure 6F). The proportion of H-zone actin (i.e., hypercontracted) sarcomeres was not significantly different between ex15e and Zasp52-PR rescues as expected. Interestingly, no hypercontracted sarcomeres were observed upon re-expression of Zasp52-PF again indicating a successful rescue (Figure 6G). Finally, we tested flight ability and observed a significant increase in flight strength in Zasp52-PF rescues compared to Zasp52-PR rescues, with the latter being not significantly different from ex15e (Figure 6H). Zasp52-PF rescues were still weaker fliers than control flies suggesting a partial rescue of flight ability. Overall, only a transgene containing exon 15e resulted in rescue, showing that the phenotypes are due to the IDR.

ex15e defects are restored by rescue constructs.

Zasp52-PF or Zasp52-PR was heterozygously overexpressed with a UH3-GAL4 driver in a homozygous ex15e background in a rescue attempt. In this figure, all myofibrils had actin visualized with phalloidin in green. All experiments were done at three weeks of age, and ctrl and ex15e data are same as previous. (A) Schematic showing Zasp52-PF which is the canonical full-length isoform, and Zasp52-PR which contains all structured domains but lacks exon 15e. PDZ domain (blue), ZM (red), LIM domains (green). (B-E) Myofibrils display bending, Z-disc disruption, and overall myofibrillar disorganization in Zasp52-PR overexpressions (D), similar to ex15e (C), while Zasp52-PF overexpression (E) appears similar to ctrl (B). (F) Bend angles were equivalent (≡) between ex15e and ex15e; UH3>Zasp52-PR, and between ctrl and ex15e; UH3>Zasp52-PF. Sample sizes: ex15e; UH3>Zasp52-PR n=2049 sarcomeres, ex15e; UH3>Zasp52-PF n=2221 (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; ex15e-ex15e; UH3>Zasp52-PR tL=1.92, tU=-10.16; ctrl-ex15e; UH3>Zasp52-PF tL=17.65, tU=-8.51). (G) The proportion of sarcomeres that were contracted (i.e., actin in H-zone) was not significantly different between ex15e and ex15e-ex15e; UH3>Zasp52-PR, whereas no such sarcomeres were found in ex15e; UH3>Zasp52-PF. Sample sizes: ex15e; UH3>Zasp52-PR n=2843 sarcomeres, ex15e; UH3>Zasp52-PF n=2480 (Fisher’s exact test; p=0.0521). (H) Flight ability was significantly improved in ex15e; UH3>Zasp52-PF (n=85) compared to ex15e; UH3>Zasp52-PR (n=66), with ex15e; UH3>Zasp52-PR showing no significant difference from ex15e but ex15e; UH3>Zasp52-PF showing significant difference from ctrl. Red arrows indicate the average position landed in ctrl flies, yellow are ex15e, green are ex15e; UH3>Zasp52-PR, and blue are ex15e; UH3>Zasp52-PF (Fisher’s exact test; ex15e; UH3>Zasp52-PR-ex15e; UH3>Zasp52-PF p<0.0001; ex15e-ex15e; UH3>Zasp52-PR p=0.2668; ctrl-ex15e; UH3>Zasp52-PF p=0.0129).

ex15e and actin genetically interact

We next wanted to test for a genetic interaction between exon 15e and Act88F (the actin isoform present in IFM) by analyzing heterozygous ex15e, heterozygous Act88FKM88 (an amorphic mutant of Act88F), and a di-heterozygous combination of both (ex15e/+; Act88FKM88/+) (Figure 7A-D). ex15e/+ resembled control myofibrils and Act88FKM88/+ has mild myofibrillar defects. ex15e/+; Act88FKM88/+ exhibited supra-additive effects where we often noticed a novel phenotype of very large H-zones. A ratio of twice the thin filament length to the total sarcomere length was computed for all genotypes. Lower values represented sarcomeres whose thin filaments occupied a smaller proportion of the sarcomere and consequently had larger H-zones, whereas a value of one represented the absence of an H-zone. This ratio was not significantly different between control, ex15e/+, and Act88FKM88/+, but was considerably smaller in di-heterozygotes (Figure 7E), indicating a strong genetic interaction between Act88F and exon 15e.

Large H-zones demonstrate a genetic interaction between exon 15e and Act88F.

(A-D) Myofibrils of one-day-old flies of various genotypes. Actin was visualized with phalloidin in green. Ctrl (A) and heterozygous ex15e myofibrils (B) appear indifferentiable; heterozygous Act88FKM88 (C) are slightly disrupted with fraying and slight narrowing. Representative ex15e/+; Act88FKM88/+ myofibrils (D) with large H-zones are shown. (E) We quantified the ratio of twice the thin filament length denoted distance “A” to the corresponding sarcomere’s length denoted as “B” for all four genotypes. A schematic illustrates the measurement of these distances. Sample sizes were as follows: ctrl n=119 sarcomeres, ex15e/+ n=122, Act88FKM88/+ n=119, ex15e/+; Act88FKM88/+ n=119. The ratio A/B represents the proportion of the sarcomere length occupied by thin filaments and is inversely correlated with H-zone length; ex15e/+; Act88FKM88/+ sarcomeres have significantly larger H-zones than all others (unpaired t-test with Welch’s correction; ctrl-ex15e/+ p=0.0502; ctrl-Act88FKM88/+ p=0.2770; ctrl-ex15e/+; Act88FKM88/+ p<0.0001; ex15e/+-Act88FKM88 p=0.4938; ex15e/+-ex15e/+; Act88FKM88 p<0.0001; Act88FKM88/+- ex15e/+; Act88FKM88 p<0.0001). All sarcomeres with an Act88FKM88 allele were radially narrower than ctrl, but a synthetic enhancement could not be concluded from this observation.

FRAP reveals defects in ex15e protein dynamics

We decided to test whether exon 15e could localize to the Z-disc on its own. To do this, we expressed the GFP-tagged exon 15e using the UH3-GAL4 driver (UH3>GFPexon15e). We observed successful localization to the IFM Z-discs (Figure 8A) and thus decided to assess kinetics using Fluorescence Recovery After Photobleaching (FRAP). One-day-old IFM Z-discs were photobleached and recovery of fluorescence was monitored. Compared to GFP-tagged Zasp52-PR (UH3>GFPZasp52-PR) which lacks exon 15e, exon15e-GFP displayed much lower recovery over a 96 second period (Figure 8B). This suggests that the IDR stabilizes Zasp52 at the Z-disc and reduces its turnover. We then expressed GFP-tagged Zasp52-PR, which cannot rescue (see Figure 6), in an ex15e background (ex15e; UH3>GFPZasp52-PR). Fluorescence recovery in the ex15e background was to a significantly higher level compared to the wild type background (Figure 8C), resulting in a larger mobile fraction (Figure 8D). A greater mobile fraction in our mutant background confirms that the IDR is required to hold Zasp52 in place at the Z-disc, either by fastening itself, and/or keeping other molecules in place thereby contributing to overall Z-disc stability.

FRAP reveals defects in ex15e protein dynamics.

(A) GFP-tagged exon 15e localized to the IFM Z-discs in one-day-old females. (B) FRAP experiments visualizing recovery of either GFP-tagged exon 15e (UH3>GFPexon15e, n=8) or GFPZasp52-PR (UH3>GFPZasp52-PR, n=12) were performed on one-day-old female IFM Z-discs. (C) Recovery of GFPZasp52-PR was also assessed in one-day-old female IFM Z-discs either in our mutant background (ex15e;UH3>GFPZasp52-PR, n=9 flies) or wild-type background (same data as in B). Ctrl UH3>GFPZasp52-PR Z-discs fixed with 4% paraformaldehyde for 10 minutes (labelled “Fixed”, n=7) displayed no recovery as expected. (D) Plateaus of fitted curves were significantly different, with a greater mobile fraction in the mutant background (unpaired t-test with Welch’s correction; p<0.0001). (E) Representative snapshots of bleached Z-disc recovery illustrate this difference between genotypes.

ex15e defects are rescued by immobilization

On a different note, we noticed that virtually all the ex15e phenotypes worsened with age, which suggests that use might contribute to deterioration. We therefore decided to prevent flies from using their IFMs by immobilizing them. Flies were reared until three weeks of age in between glass plates separated by 1.5 mm spacers which prevented them from flying or even attempting to fly. Myofibrils appear normal in immobilized ex15e mutant flies (Figure 9A-C). The degree of bending in immobilized flies was equivalent to that of control flies (Figure 9D). The proportion of hypercontracted sarcomeres was also not significantly different between immobilized and control flies (Figure 9E). Intriguingly, even flight ability was restored in immobilized flies compared to age-matched controls with no significant difference between the two (Figure 9F). These results suggest successful rescue by preventing IFM use via immobilization.

ex15e defects are rescued by immobilization.

Rescues were attempted by immobilizing flies thereby preventing IFM use. ex15e myofibrils in flies that were immobilized (C) resembled those of ctrl (A), both showing no phenotypes compared to ex15e (B). (D) Bend angles were equivalent between ctrl and ex15e immobilized flies (n=2119) (Two One-Sided Test (TOST) with ±4.5° equivalence margin, α=0.05; tL=15.36, tU=-12.00). (E) The proportion of sarcomeres that were contracted was not significantly different between ctrl and immobilized flies (n=2246) (Fisher’s exact test; p=0.9037). (F) Flight ability was rescued with immobilized flies (n=67 flies) having no significant flight defect compared to ctrl (n=86) but being significantly improved from their non-immobilized counterparts (n=81) (Fisher’s exact test; ctrl-immobilized p=0.0640; ctrl-ex15e and ex15e-immobilized p<0.0001).

Discussion

Traditionally, IDRs were associated with the concept of amorphousness, finding themselves in transient structures such as condensates and membraneless organelles. As our understanding of disordered proteins has changed, their role in more disparate contexts has become increasingly appreciated, revealing them as modular hubs, flexible scaffolds, and more (Cortese et al., 2008). Furthermore, their role in disease has spurned great interest in their myriad functions. In this study, we demonstrate the role of an exceptionally large disordered region, Zasp52’s exon 15e, in maintaining stability and integrity in a highly rigid and force-bearing structure – the Z-disc.

First, we showed that the expression of exon 15e is spatiotemporally regulated. Being confined to the adult IFM suggests that exon 15e’s function is uniquely required for the rapid oscillatory contractions in indirect flight. Supporting this is the fact that the linker region containing exon 15e is generally much longer in Drosophilids, and rather short in insects using direct flight. Although this linker region is predicted by multiple software programs to be disordered, within it are regions of greater order, one of which corresponds to a putative actin-binding motif (Ashour et al., 2023). This arrangement exists in other proteins such as Cno in Drosophila, where higher-ordered actin-binding regions are embedded in a large IDR (Jensen et al., 2025). Furthermore, a synthetic enhancement between ex15e and Act88FKM88 heterozygotes indicates that exon 15e and actin genetically interact. There may be other important short linear motifs (SLiMs) contained within the linker region which serve as interacting sites with other sarcomere proteins. Indeed, multiple sequence alignment of the linker regions of various insects reveals several conserved motifs of unknown function. The presence of an actin-binding motif that may strengthen thin filament anchoring to the Z-disc would be in line with the phenotypes we observed in ex15e. However, we were unable to detect defects in our deletion of this motif which supports the theory that the entire IDR is essential and that the ABM is not the key to its function. It is still possible that the ABM within exon 15e acts in concert with the N-terminal ABM, but specific amino acids mediating the N-terminal actin binding have not yet been identified (Liao et al., 2020). Bending at the Z-disc is indicative of a loss of thin filament anchoring and stability at the Z-disc, allowing them to shift out of alignment. Normal sarcomeres bent due to force would exhibit a kinked-hose appearance, where the inner edge is sharply bent and the outer one bowed. This was occasionally observed in controls, whereas ex15e bends were always shifted uniformly. Actin in the H-zone is also indicative of loss of stable thin-filament anchoring. During preparation, IFMs are incubated overnight in skinning solution during which they enter a state of rigor. After washing them with an ATP-containing relaxing solution, the myosin heads detach from the thin filaments and the supple nature of the sarcomere which provides it with elastic recoil, causes it to resume a relaxed conformation. However, when thin filaments are weakly anchored at the Z-discs, the natural tension contained within the sarcomere is reduced and thus it will not “bounce back” into a relaxed conformation, resulting in a hypercontracted state which manifests as actin continuing to reside in the H-zone. Cheerio, a Drosophila filamin, was shown to stably attach actin thin filaments at the Z-disc and induces a similar actin incorporation into the H-zone when mutated (González-Morales et al., 2017). Thus, we also demonstrate the importance of Z-disc integrity in allowing sarcomeres to return to their relaxed state in rapidly contracting asynchronous IFM. Accordingly, synchronous muscles do not have such regularly structured Z-discs, and do not require Zasp52’s exon 15e. Indeed, exon 15e is absent in larval and other muscle types. We therefore propose that exon 15e is uniquely required to stabilize rapidly contracting asynchronous IFM.

The orthologous human ALP/Enigma family consists of seven members, with LDB3 (ZASP) considered the canonical Zasp52 ortholog. Zasp52 is more precisely a combined ALP/Enigma member; its first LIM domain is sequentially homologous to the single LIM domain of human ALP family members, and its last three LIM domains are homologous to the three Enigma family LIM domains. Though these human proteins do contain disordered linkers connecting their structured domains, they are much shorter than in Drosophila and are not sequentially homologous. The length of Drosophila’s linker was likely evolutionarily driven to be large, as we see insects using direct flight to have much shorter linkers. Several pathological mutations have been identified in the linker region between LDB3’s PDZ and LIM domains (Vatta et al., 2003), some of which are now known to lie in the actin-binding Zasp motif (ZM) (Lin et al., 2014), which corresponds to the actin-binding region N-terminal to Zasp52 LIM1 (Liao et al., 2020). Interestingly, LDB3 has three skeletal muscle-specific splice isoforms which are differentially expressed and variably retain/omit exons 9 through 11 of a linker sequence (Lin et al., 2014). The prenatal long (L) isoform contains exons 10 and 11, post-natal long (LΔex10) has only exon 11, and short (S) has only 9. Post-natal LDB3-LΔex10 contains an actin-binding region in its linker region whose function is abolished by the inclusion of exon 10 in LDB3-L. Thus, human Zasp protein linkers also function in actin binding, regulated by developmentally differential expression. Linker regions in several human ALP/Enigma proteins were found to interact with both α-actinin and actin using AlphaFold (Healy & Collins, 2023). The mouse ortholog Cypher also contains tissue-specific splice variants which variably include or omit linker sequences (Huang et al., 2003). Like Zasp52’s exon 15e, Cypher is not required for sarcomerogenesis or Z-disc assembly but only maintenance (Zhou et al., 2001).

Our genetic rescues support our hypothesis of exon 15e involvement in stabilizing thin filaments and Z-discs. The rescues also demonstrate that the conserved domains (PDZ, ZM, LIM1-4) are not sufficient for proper Z-disc and thin filament stability in IFM, because Zasp52-PR could not rescue any of the defects. Interestingly, these phenotypes, as well as the flight defects we observed, worsened with age, with defects in one-day-old flies being very minor. This implies that exon 15e is not required developmentally, but rather for the maintenance of sarcomeres which naturally deteriorate with use. We originally hypothesized that since IDRs often lead to protein aggregation and phase separation, exon 15e might drive the formation of Z-bodies, primitive Z-discs which form through the coalescence of diffuse Z-disc proteins including Zasp52. However, given the lack of defects in early Z-discs and the fact that exon 15e expression does not temporally coincide with Z-body formation, we reject this hypothesis. The role of exon 15e in maintenance is supported by immobilization rescues, in which flies were prevented from flying. We opted for a low-ceiling apparatus which we believe would not only prevent flies from flying but discourage them from futilely attempting to utilize their IFM at all. Surprisingly, immobilization restored flight ability to wild type levels. Interestingly, human zaspopathies often manifest late in life, with one study citing onset at 44 to 73 years of age (Selcen & Engel, 2005), much like what we observed in flies. Our data suggest that avoiding strenuous exercise might delay the onset of zaspopathies. Given the maintenance-specific role of exon 15e, but also that as a whole, Zasp52 is indispensable for pupal myofibril development and not just maintenance (Katzemich et al., 2013, Medina-Quintana et al., 2026), Zasp52 appears to have a multifaceted role more complex than previously considered.

It was demonstrated in Liao et al., 2016 that various structured domains of Zasp52, such as the C-terminal LIM domains, are sufficient to localize to the Z-disc. We found that the disordered exon 15e was also able to localize to the Z-disc indicating that some feature contained within it is capable of targeting the protein to its correct location. We took advantage of this to assess its kinetics using FRAP and found that the exon 15e construct is much less dynamic than Zasp52-PR (which lacks exon 15e) suggesting that this disordered region is important for anchoring the protein in place. This finding supports the theory that the IDR acts to stabilize thin filaments at the Z-disc. Although one-day-old ex15e myofibrils are morphologically indistinguishable from controls, we were already able to detect abnormalities in Zasp52 dynamics using FRAP in our mutant background. The presence of recovery suggests that certain Zasp52 isoforms are not immobile as might be assumed for an anchoring protein and do indeed possess kinetics on a sub-minute timeframe. This is supported by significant transcription of the gene, including exon 15e-containing isoforms, into adulthood. The mobile fraction of Zasp52-PR was significantly increased in a system entirely lacking exon 15e (the PR isoform also lacks exon 15e) indicating a role of exon 15e in retaining Zasp52 at the Z-disc, whether by direct interaction or not. The FRAP data suggest that Zasp52’s IDR stabilizes the Z-disc, in line with the electron microscopy data. It suggests that Z-discs can be described as highly solid aggregates in part stabilized by IDRs, on the other end of the spectrum of the transient condensates initially described, thereby further expanding the importance of IDRs for establishing subcellular structures.

Overall, Zasp52 and its isoforms represent a complex toolbox of multifunctional proteins. There has been evidence suggesting that Zasp52 is not required for initial sarcomere assembly in embryos (Rui et al., 2010), important for Z-disc maintenance overall (Rui et al., 2010, Jani & Schöck, 2007, Katzemich et al., 2013), and required for pupal myofibril assembly (Katzemich et al., 2013). The complete picture of Zasp52’s varied function is still unclear; however, we have shown that a specific disordered region is required for maintaining IFM and stabilizing the rigidity of Z-discs. The long linker might not only contain protein-interacting regions and provide the glue to hold the Z-disc together but might also provide the spacing needed for the neighboring structured domains to properly interact with their partners and perform their roles. We have seen in our genetic interaction experiment an inability of thin filaments to properly elongate when Act88F is partially depleted, which further implicates exon 15e not only in Act88F structure but dynamics. Most encouraging is the complete rescue of flight ability through immobilization, which may provide avenues for delaying the onset of human zaspopathies through reducing strenuous exercise.

Materials and Methods

Fly strains and husbandry

Flies were housed at room temperature (22°C) in standard glass vials and food unless indicated otherwise. Immobilized flies were reared in between 3.25×3.25 inch glass plates separated by 1.5 mm spacers; standard food was inserted into the gap. Immobilized flies were transferred to a new apparatus with fresh food every couple of days. ex15e flies were created by Genetivision using CRISPR-mediated homology-directed repair. The gRNA sequences used were TTTGAACTTGAACTCTAATGTGG for the downstream target and CAGCCTCTCCACCTTTCCGATGG for upstream. The deletion was replaced with a cassette containing a 3xP3DsRed marker flanked by attP landing sites. It was validated by sequencing at the cut sites. ΔABM flies were created by Genome ProLab using CRISPR-mediated homology-directed repair as well. The gRNA sequence GAGGAGCTCTCGGGTCTGCA was used to precisely delete five amino acids (LQREL), verified by sequencing. The following genotypes were used: ex15e, ΔABM, w1118 as control in all experiments, Df(2R)BSC427, Act88FKM88, UAS-GFPZasp52-PR, UH3-Gal4, UAS-Zasp52-PF, UAS-Zasp52-PR, UAS-GFPexon15e, and Zasp52MI00979. UAS-Zasp52-PF and UAS-Zasp52-PR consist of the respective Zasp52 isoform plus an N-terminal ALFA-tag and C-terminal Spot tag. UAS-GFPexon15e encompasses all amino acids encoded by exon 15e except the last 9 encoding the beginning of LIM2. All UAS constructs were made by Genscript and injected into w1118 by GenomeProlab.

Western blotting

Extracts were prepared by homogenizing ten flies/fly parts in 100 µL RIPA buffer with protease inhibitor plus an equal amount of 2× Laemmli buffer with β-mercaptoethanol. Samples were boiled at 95°C for five minutes and centrifuged at 12,000×rcf for 5 minutes. Supernatant was used for electrophoresis according to standard protocol using a 4-15% BioRad TGX Stain-free gel. Stain-free gels were imaged for total protein content on a BioRad ChemiDoc before protein was transferred onto a nitrocellulose membrane. Standard Western blotting procedures were followed. Primary antibodies used were rabbit anti-Zasp52 full-length (Jani & Schöck, 2007), and a rabbit antibody against the three C-terminal LIM domains of Zasp52 (FEKYLAPTCSKCAGKIKGDCLNAIGKHFHPECFTCGQCGKIFGNRPFFLEDGN AYCEADWNELFTTKCFACGFPVEAGDRWVEALNHNYHSQCFNCTFCKQNLEGQSFYNKGGRPF CKNHAR) made by Genscript; secondary antibody was goat anti-rabbit IgG (H+L) HRP (ThermoFisher Scientific).

Locomotory assays

For flight assays, flies were reared in vials (unless immobilized) until the desired age, with vials flipped whenever pupae began to appear. Flight assays were conducted in a manner similar to that described in Babcock & Ganetzky, 2014. A tube 22 cm long and 3.2 cm in diameter was placed in a funnel; the bottom of the tube rested 4 cm above the bottom spout of the funnel. The funnel rested in a 60 cm long tube which was 13.5 cm in diameter and covered with TAD Shake & Spray glue (LADD Research) on the inside. The bottom of the funnel was 12 cm below the top of the larger tube. The larger tube was divided vertically into seven 8.5 cm segments, with segment 1 being at the top, and segment 7 being at the bottom. A dish containing water was placed below the tube and was considered segment 8. Vials containing flies were dropped down the narrow tube so that they would be ejected into the larger tube and land somewhere along its length. Flies naturally fly upwards and those with stronger flight would land higher up the tube. Binary flight assays were conducted by releasing flies one at a time and recording whether they could generate any upwards lift. They were raised in the same conditions as above.

For larval crawling assays, third instar larvae of the desired genotype were collected from bottles and assayed according to Post & Paululat, 2018. In brief, videos of larvae on a slightly moistened petri dish were taken. Time segments where larvae were crawling in a straight line continuously for at least ten seconds were analyzed – contractions per distance were measured and recorded.

Muscle preparation and immunofluorescence

Primary antibodies used were rabbit anti-Zasp52 full-length (Jani & Schöck, 2007) and mouse anti-βPS-integrin (CF.6G11; obtained from Developmental Studies Hybridoma Bank, Brower et al., 1984) at a concentration of 1:200 for both. Secondary antibodies used were Alexa Fluor 647 goat anti-rat (ThermoFisher Scientific) and Alexa Fluor 488 goat anti-rabbit (ThermoFisher Scientific) at a concentration of 1:400.

For IFM tissue, a protocol similar to Xiao et al., 2017 was followed (refer to for recipes). Briefly, thoraces with bisected and incubated in Relaxing-Glycerol solution at -20°C overnight. IFM were isolated and fixed in 4% paraformaldehyde for 15 minutes. Fixed IFM were incubated in primary antibodies at 4°C overnight. IFM were incubated in Acti-stain 555 fluorescent phalloidin (Cytoskeleton Inc) and secondary antibodies, at room temperature for one hour. Samples were mounted in Mowiol mounting medium.

For thorax hemisections including MTZ, a scalpel was used to sagittally bisect thoraces. Samples were fixed in 4% paraformaldehyde for 15 minutes and incubated in primary antibodies for one hour at room temperature, and then Acti-stain 555 fluorescent phalloidin and secondary antibodies for one hour at room temperature. Samples were mounted in Mowiol mounting medium using three 0.12 mm spacers.

For TEM, bisected thoraces were treated with 5 mM MOPS, 150 mM KCl, 5 mM EGTA, 5 mM MgCl2, 5 mM ATP, 1% Triton X-100, and protease inhibitor for two hours. then the same solution with 50% glycerol instead of Triton X-100 overnight at 4°C. This was repeated at -20°C. Samples were washed in 5 mM MOPS, 40 mM KCl, 5 mM EGTA, 5 mM MgCl2, 5 mM NaN3, protease inhibitor, and then fixed in 3% glutaraldehyde and 0.2% tannic acid for two hours. Samples were washed in 20 mM MOPS, 5 mM EGTA, 5 mM MgCl2, 5 mM NaN3, and then in 100 mM sodium phosphate buffer, 5 mM MgCl2, 5 mM NaN3. Secondary fixation was performed in 1% osmium tetroxide in the previous sodium phosphate buffer for one hour. Samples were stained in 2% uranyl acetate, washed with water, and underwent a serial dehydration in ethanol. Samples were washed with propylene oxide and serially infiltrated with Mollenhauer Epon-Araldite. After curing in Beem capsules, 70 nm sections were cut and stained in 4% uranyl acetate followed by SATO lead stain.

Fluorescence Recovery After Photobleaching

The sequence encoding exon 15e minus the nine C-terminal amino acids encoding LIM2 was subcloned into pUAST-attB-GFP-V5-His and injected into w1118. IFM were prepared as previously described (with fixation) and stained with ActiStain-555 and FluoTag®-X4 anti-GFP (NanoTag Biotechnologies) to visualize localization; FRAP was performed as described below. One-day-old female flies were anaesthetized and their abdomens, wings, and legs removed. Carcasses were pinned down by the head on a silicone dish and immersed in hemolymph-like 3 (HL3) saline (Stewart et al., 1994). Thoraces were sagittally bisected, washed thrice in HL3 saline, and mounted with a 0.12 mm spacer. Imaging was performed immediately using a HC PL APO 63×/1.40 OIL CS2 objective on a Leica SP8 point-scanning confocal microscope. Five pre-bleach frames (1.29 seconds each) were recorded before bleaching with a 488-nm laser to achieve a bleach depth averaging 75%±4% (SD) for UH3>GFPexon15e, 74%±8% for UH3>GFPZasp52-PR, 77%±7% for ex15e; UH3>GFPZasp52-PR, and 76%±9% for fixed samples. Seventy-five post-bleach frames (1.29 seconds each) were then acquired. ROIs (1.25×3 µm) were placed over individual Z-discs and those with drift were discarded. For each fly, three Z-discs on different myofibrils were bleached, and intensities of adjacent unbleached Z-discs and unbleached background regions were recorded. Fluorescence was normalized as (bleached–background)/(unbleached–background), averaged across the three ROIs, and scaled such that the post-bleach frame equaled 0 and the pre-bleach mean equaled 100. Recovery curves were presented as mean±SEM. Each fly’s curve was fitted to a one-phase exponential association model to obtain the plateau (mobile fraction). Fixed UH3>GFPZasp52-PR samples were prepared in the same manner except with an additional ten-minute fixation in 4% paraformaldehyde in HL3 saline.

Microscopy

Confocal images were acquired using a HC PL APO 63×/1.40 OIL CS2 objective on a Leica SP8 point-scanning confocal microscope. The position along the z-axis was adjusted until perfectly in the centre of the myofibril, i.e., the widest point. Polarized light images were acquired using a 5X N-ACHROPLAN, NA=0.13 objective on a Zeiss Axio Observer microscope. TEM images were acquired using an FEI Tecnai G2 Spirit Twin 120 kV Cryo-TEM using an AMT NanoSprint15 MK2 CMOS camera.

Computational analyses

Schematics were illustrated with IBS 2.0 (Xie et al., 2022). All protein domains were defined per SMART (Letunic & Bork, 2018). Graphing and statistical analysis was performed using GraphPad Prism 10. Statistical analyses were performed in GraphPad Prism 10, where ns=p≥0.05, *=p<0.05; **=p<0.01, and ****=p<0.0001, and calculated to four decimal places. Equivalence of bend angles was assessed using the Two One-Sided Test (TOST) procedure. Model sarcomeres with bend angles ranging from 180° (straight) to 95° and blurring were generated. Each was hand measured ten times and the average absolute measurement error (4.5°) was used as the equivalence margin for all TOST analyses.

Minus strand RNA-seq bigWig files from the ENCODE project (Duff et al., 2015) with the following accessions were retrieved and loaded in IGV 2.19.4 (Robinson et al., 2011): ENCFF289ZZT, ENCFF982OGI, ENCFF430HWM, ENCFF716DNU, ENCFF829OGQ, ENCFF936YJM, ENCFF987YVI, ENCFF033MIY, ENCFF845HII, ENCFF021VXY, ENCFF792FNT. Signal tracks were aligned to the dm3 assembly and the data range was adjusted uniformly for all tracks.

SRA files with the following accession numbers were downloaded from the Bioproject PRJNA419412 (Spletter et al., 2018): SRR6314253, SRR6314256, SRR6314275, SRR6314259, SRR6314277, SRR6314262, SRR6314273. Salmon 1.10.0 (Patro et al., 2017) was used to quantify expression of the following transcripts which correspond to identified Zasp52 isoforms: FBtr0392902, FBtr0329914, FBtr0329912, FBtr0100387, FBtr0111048, FBtr0087315, FBtr0329913, FBtr0310086, FBtr0329916, FBtr0329915, FBtr0310088, FBtr0310087, FBtr0392901, FBtr0329911, FBtr0342711, FBtr0100388, FBtr0302163, FBtr0301313, FBtr0479884, FBtr0329909, FBtr0392918, FBtr0329910. The first six wereclassified as containing only exon 15a, the next six containing exons 15b-d, the next four containing exon 15e, and the last six containing no exon 15.

ImageJ was used to evaluate bending manually using the angle tool, and sarcomere length using the line measuring tool.

For disorder prediction, the entire Zasp52-PF sequence was entered into AIUPred v0.1 (Erdős & Dosztányi, 2024), DISOPRED3 (Jones & Cozzetto, 2015) (as two sequences, the first covering the first 1200 amino acids and the second, the remaining), or metapredict online v3.0 (Emenecker et al., 2021). Raw values for each residue were obtained and smoothed using a Savitzky-Golay filter with fourth order smoothing and 11 neighbours on each side. AlphaFold v2.0 (Jumper et al., 2021) was used to obtain pLDDT scores for the alpha carbon of each residue. pLDDT scores were min-max normalized and smoothed in the same manner as above.

Zasp52 orthologs were obtained for members of the class Insecta using NCBI’s Eukaryotic Genome Annotation Pipeline. Canonical protein sequences for each were downloaded and those which did not contain the classical PDZ and four LIM domains were removed. The number of residues between the first and second LIM domains was calculated. A phylogenetic tree of these species was created using NCBI’s Common Tree (Schoch et al., 2020) and a heatmap created using iTOL v.6 (Letunic & Bork, 2024).

NARDINI (Cohan et al., 2022) was used for the sequences indicated to generate a z-score matrix, with 100,000 scrambled sequence variants.